---
title: "HMM_behaviors"
author: "Camila Calderon"
date: "2024-08-05"
output: html_document
---

### Load libraries

```{r setup}
library(momentuHMM)
library(tidyverse)
library(lubridate)
library(parallel)
library(mixtools)
library(plotly)
library(ggmap)
library(adehabitatLT)
```

### HMM for all data where GPS sampling is every 2 minutes 

```{r, results=FALSE}
# load data
load(file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/data/clean_bocasbats.RData")

# if step and angle are present need to be removed
names(bocas)
bocas <- bocas[!names(bocas) %in% c("step", "angle")]

# order by timestamp and ID
bocas <- arrange(bocas, ID_batday,timestamp)

# create ADE trajectory added with NAs every 2 minutes for all individuals. I decided not to interpolate as it was creating unrealistic tracks. 

# add NA's every 2 minutes
df_na <- setNA(ltraj = as.ltraj(xy=bocas[, c("x", "y")], date = bocas$timestamp, id=bocas$ID_batday), date.ref =bocas$timestamp[1], dt=2, tol=2, units="min")
# convert ltraj object to dataframe
df <- ld(df_na)[,c("x","y", "id","date")]

# split data frame by idday
bats.split <- split(df, df$id)

# process data with Prepdata
batsData<- lapply(bats.split, function(x){
  momentuHMM::prepData(x, type = 'UTM', coordNames = c('x', 'y'))
})

# rename the id column with right id
batsData <- lapply(1:length(batsData), function(i){
  id_day <- names(batsData[i])
  batsData[[i]]$ID <- id_day
  return(batsData[[i]])
})

# to check steps one id
hist(batsData[[40]]$step)

# plotting to see step lengths and turning angles
pdf(file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/HMM/Parameters.pdf")
lapply(batsData, function(x){
  tryCatch({
    plot(x, ask=FALSE)
  },error=function(e) finally = print("f'd up"))
})
dev.off()
```

### Classify behaviors: foraging and commuting with a HMM 2 state model

```{r pressure, results=FALSE}
# estimate parameters for HMM
batsData.df <- do.call(rbind, batsData)

# calculates normal distribution from data and select 2 values from it
mixHMM2 <- normalmixEM(na.omit(batsData.df$step), k = 2)
# plot(mixHMM,which=3)
idx2 <- order(mixHMM2$mu)
# select step length values
mu02 <- c(mixHMM2$mu[idx2[1]], mixHMM2$mu[idx2[2]])
# select step SD
sigma02 <- c(mixHMM2$sigma[idx2[1]],  mixHMM2$sigma[idx2[2]]) # step SD

## Fitting HMM with 2 states
state2Names <- c("state1", "state2")
# initial parameters
step2ParMean <- mu02
step2ParSD <- sigma02
zero2mass0 <- c(0.01, 0.01)
step2Par0 <- c(step2ParMean,step2ParSD,zero2mass0)#zero2mass0
angle2Par0 <- c(0.5,0.1)
dist = list(step = "gamma", angle = "wrpcauchy")
m <- fitHMM(data=batsData.df, 
            nbStates=2, 
            dist=dist,
            Par0=list(step=step2Par0, angle=angle2Par0),
            stateNames = state2Names)

# save results 
pdf(file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/HMM/HMM_2states.pdf")
plot(m, plotCI = TRUE, ask=FALSE)
dev.off()
```

### Checking classification from the model and add it with the original data frame
```{r, results=FALSE}
# viterbi function reconstructs most probably states sequence for each bat
m$data$HMM2 <- viterbi(m) 
m$data$HMM2_prob <- stateProbs(m)

# order data again
bocas <- bocas[order(bocas$ID_batday, bocas$timestamp),]

# remove the NAs that were included in the beggining
m2_df <- m$data[which(!is.na(m$data$x)),]

# adding the states to the initial data frame
bocas$HMM <- m2_df$HMM2[match(bocas$timestamp, m2_df$date)]
bocas$HMMprob <- m2_df$HMM2_prob[match(bocas$timestamp, m2_df$date)]
bocas$turningangle <- m2_df$angle[match(bocas$timestamp, m2_df$date)]
bocas$step <- m2_df$step[match(bocas$timestamp, m2_df$date)]
```

## Change classification points which are wrongly assigned in either of the behaviors

```{r, results=FALSE}
# change state numbers to behaviors and adjust manually some of the classification
bocas$behav <- bocas$HMM
bocas$behav[which(bocas$HMM==2)] <- "commuting"
unique(bocas$behav)
bocas[which(is.na(bocas$behav)),]
bocas$behav[which(bocas$HMM==1)] <- "foraging"
bocas$behav[(bocas$behav=="commuting" & bocas$ground_speed<1)] <- "foraging"
# some points that are flying are classified as foraging
bocas$behav[(bocas$behav=="foraging" & bocas$ground_speed>3)] <- "commuting"

# checking for na's
bocas <- bocas[which(!is.na(bocas$behav)),]

# plot foraging points results
mapview::mapView(bocas[bocas$behav=="foraging",], xcol="location_long", ycol="location_lat", zcol="behav", legend=F, crs="EPSG:4326") #as points

# plot foraging points results
mapview::mapView(bocas, xcol="location_long", ycol="location_lat", zcol="behav", legend=F, crs="EPSG:4326") #as points

# correcting points that are in the water
bocas$behav[(bocas$event_id=="22068516465" |bocas$event_id== "22068504065" | bocas$event_id== "29640513640"|bocas$event_id=="29640503867" | bocas$event_id=="22068516452" | bocas$event_id=="22068505980" | bocas$event_id=="29640589130" | bocas$event_id=="29640585899" | bocas$event_id=="29640511978" | bocas$event_id=="1184403553" | bocas$event_id== "22068512106" | bocas$event_id== "22068515511" | bocas$event_id=="22068514769" )] <- "commuting"
 
	#1184403553  	22068512106 22068515511 22068514769 	 	 		  
	#22068504065 22068516465 29640513640 29640503867 22068516452 22068505980 29640589130 29640585899 29640511978
		
# look at the histograms of speed and turning angle
hist(bocas[bocas$behav=="commuting",]$ground_speed)
hist(bocas[bocas$behav=="foraging",]$ground_speed)
hist(bocas[bocas$behav=="commuting",]$turningangle)
hist(bocas[bocas$behav=="foraging",]$turningangle)

# save RData
save(bocas,file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/data/bocas_2022-2023HMMbehaviors.RData")
```


## Classify behaviors: foraging and commuting with a HMM 3 state model
```{r, results=FALSE}
# calculates normal distribution from data and select 3 values from it
mixHMM3 <- normalmixEM(na.omit(batsData.df$step), k = 3)
# plot(mixHMM,which=3)
idx <- order(mixHMM3$mu)
# select step length values for the three states
mu03 <- c(mixHMM3$mu[idx[1]], mixHMM3$mu[idx[2]], mixHMM3$mu[idx[3]])
# select step SD for the three states
sigma03 <- c(mixHMM3$sigma[idx[1]],  mixHMM3$sigma[idx[2]],mixHMM3$sigma[idx[3]]) # step SD

# look at what distribution to use
library(fitdistrplus)
# step length
plotdist(batsData.df$step[!is.na(batsData.df$step)], histo = TRUE, demp = TRUE)
descdist(batsData.df$step[!is.na(batsData.df$step)], boot = 1000, method="unbiased")#beta
# turning angle
plotdist(as.numeric(batsData.df$angle[!is.na(batsData.df$angle)]), histo = TRUE, demp = TRUE)
descdist(as.numeric(batsData.df$angle[!is.na(batsData.df$angle)]), boot = 1000, method="unbiased")

# fitting HMM for whole individuals with 3 states
state3Names <- c("state 1","state 2", "state 3")
# initial parameters
step3ParMean <- mu03
step3ParSD <- sigma03
zero3mass0 <- c(0.01, 0.01, 0.01)
step3Par0 <- c(step3ParMean,step3ParSD)#zero3mass0
angle3Par0 <- c(0.5,0.1,0.01)
dist = list(step = "gamma", angle = "wrpcauchy")
m3 <-fitHMM(data=batsData.df, 
            nbStates=3, 
            dist=dist,
            Par0=list(step=step3Par0, angle=angle3Par0),
            stateNames = state3Names)
# print model
m3

# look at classification
pdf(file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/analysis/results/HMM/HMM_3states.pdf")
plot(m3, plotCI = TRUE, ask=FALSE) 
dev.off()

# checking the two models
AIC(m, m3) # take model m3 because due to AIC is better predicted
```



