---
title: "Dynamic time warping"
output:
  html_document:
    df_print: paged
  pdf_document: default
date: "2024-03-27"
Author: Camila
---

Euclidean and DTW distances are both classified as shape based. This type takes into account the overall shape and matches time series based on that aspect. It is a whole time series analysis. #good explanation of DTW
<https://rpubs.com/esobolewska/dtw-time-series>

The main difference between both distances can be best understood graphically. The picture below shows an example of matched points of two data vectors. They were connected based on the minimal distance between points based on DTW (black) and Euclidean (red) distance. It can be seen that with DTW the 11th blue point matches 4 green points. When taking into account the Euclidean distance it can be seen that the assigned points 9-to-9 and 10-to-10 are visibly further than 9-to-11 and 10-to-11. That can significantly impact the overall distance of series between each other. The DTW takes into account the shape of both time series much better.

Dynamic Time warping is a method of calculating distance that is more accurate than Euclidean distance. It has an advantage over Euclidean if data points are shifted between each other and we want to look rather at its shape. Additionally two time series don’t have to be equal in length what is an assumption required by the Euclidean distance. The Euclidean distance takes pairs of data points and compares them to each other. DTW calculates the smallest distance between all points - this enables a one-to-many match. In literature dynamic time warping is often paired with k-medoids and hierarchical methods. In some papers it is sometimes paired with k-means though this is a controversial matter (Stack Exchange (2015)). Out of non classical methods, it was also paired with random-swap and hybrid(Aghabozorgi, Shirkhorshidi, and Wah (2015)).

```{r setup, warning=FALSE, message=FALSE}
library(dtw)  # dynamic time warping
library(sp)  # handle spatial objects
library(rgeos)# geoprocessing functions
library(lubridate)
library(dplyr)
library(ggplot2)
library(dtwclust)
library(cluster)    # clustering algorithms
library(factoextra) # clustering algorithms & visualization
library(move)
library(stringr)
library(ggmap)
library(sf)
library(purrr)
```

### Load data

```{r loading, results='hide'}
load(file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/data/bocas_2022-2023HMMbehaviors.RData")

# order data frame
bocas <- bocas[order(bocas$tag_local_identifier, bocas$timestamp),]

bocas_used <- bocas %>%
  filter(year_cave=="2022_lagruta_Feb"|year_cave== "2022_ajcave"|year_cave== "2023_ajcave")

move_path <- move(data=bocas_used,
                  x=bocas_used$location_long, 
                  y=bocas_used$location_lat, 
                  time=as.POSIXct(bocas_used$timestamp, format="%Y-%m-%d %H:%M:%S", tz="UTC"), 
                  animal = bocas_used$ID_batday,
                  sensor = bocas_used$sensor_type,
                  proj = CRS("+proj=longlat +ellps=WGS84"))

#create a movestack
# add movement parameters, pre-cleaning
moveList <- lapply(split(move_path), function(myInd){
  myInd$tlag <- c(NA, move::timeLag(myInd, units="secs"))
  return(myInd)
})

#check time tracked for each id
hourstracked <- lapply(1:length(moveList), function(id){
  #x$tlag[x$tlag>=50000] <- NA
  id <- names(moveList[id])
  s <- sum(moveList[[id]]$tlag, na.rm = TRUE)
  hours <- data.frame(idday=id, h=s/3600)
  return(hours)
})

hourstracked <- do.call(rbind.data.frame,hourstracked)
hourstracked$id <- substr(hourstracked$idday, 1,nchar(hourstracked$idday)-2)

hourssume <- hourstracked %>%
  group_by(id)%>%
  dplyr::summarise(sum_hours= sum(h), meanhours= mean(h), count = n())

phas <- moveStack(moveList, forceTz="UTC")
bocas_used <- as.data.frame(phas) 
bocas_used$tlag[which(bocas_used$behav=="foraging" & bocas_used$tlag>1800)] <- 190

samplingeffort <- bocas_used %>%
  group_by(tag_local_identifier)%>%
  dplyr::summarise(hours=sum(tlag, na.rm = TRUE)/3600)

bocas_ls <- split(bocas, bocas$ID)

```

## Select commuting tracks to the foraging spots from the colony 1 groups in February 2022

```{r, message=FALSE, warning=FALSE, results=FALSE}
# adding a variable that counts the length of consecutive commuting and foraging points. This will be used to filter only commuting tracks that contain more than 5 points in a row
bocas$n <- NA
r <- rle(bocas$behav)
i=1
for(i in 1:length(r$lengths)){
  #if(r$values[i] == "commuting"){
      start <- {}
      end <- {}
      try(start <- sum(r$lengths[1:(i-1)])+1)
      if(i == 1) start <- 1
      end <- sum(r$lengths[1:i])

      bocas$n[start:end] <- r$lengths[i]
}

# filter only commuting with less than 5 points in a row
bats_commu <- bocas %>% 
  filter(behav=="commuting" & n>=4 & location_long<=-80)

# add time difference to split commuting tracks by the ID_day
bats_commu <- bats_commu %>%
  group_by(ID_batday) %>%
  mutate(diff = timestamps - lag(timestamps))
bats_commu$diff <- as.numeric(bats_commu$diff)
bats_commu$diff[is.na(bats_commu$diff)] <- 0

# data frame with data from February 2022
commu_dry <- bats_commu %>% 
 filter(date>"2022-01-01" & date < "2022-03-01") %>%
  filter(tag_local_identifier!= "0C506E35_G")#remove individual with only 2 nights of tracking
#

# splitting tracks by ID
commu.split <- split(commu_dry, f=commu_dry$ID)

#remove the first two rows for these individuals which make the classification weird
commu.split[["2C507735_G_2022-02-01"]] <- commu.split[["2C507735_G_2022-02-01"]][c(-1,-2),]
commu.split[["2C507735_G_2022-02-03"]] <- commu.split[["2C507735_G_2022-02-03"]][c(-1,-2),]

# for loop to divide each day of the animal in commuting outbound and commuting inbound using different threshold values
mylist <- list()
for (i in 1:length(commu.split)){
  x <- which(commu.split[[i]]$diff>=600 & commu.split[[i]]$diff<=601 |commu.split[[i]]$diff>=900 & commu.split[[i]]$diff<=970 | commu.split[[i]]$diff>=1000 & commu.split[[i]]$diff<=1100 | commu.split[[i]]$diff>=1300 & commu.split[[i]]$diff<=1322 | commu.split[[i]]$diff>=1570 & commu.split[[i]]$diff<=2200 | commu.split[[i]]$diff>=2526 & commu.split[[i]]$diff<2527 | commu.split[[i]]$diff>=2402  & commu.split[[i]]$diff<=2765 | commu.split[[i]]$diff>=3123 & commu.split[[i]]$diff<=3728 | commu.split[[i]]$diff>=2800 & commu.split[[i]]$diff<=2900 | commu.split[[i]]$diff>=4000)#70 minutes in seconds  2200 2402 2523 commu.split[[i]]$diff>=482 & commu.split[[i]]$diff<=490 | 
  mylist[[i]] <- split(commu.split[[i]], cumsum(1:nrow(commu.split[[i]])%in%x))
  indv <- unique(names(commu.split[i]))
  names(mylist[[i]]) <- paste(indv, seq_along(mylist[[i]]), sep = "_")
}

# convert nested list to data frame
find_df <- function(x) {
  if (is.data.frame(x))
    return(list(x))
  if (!is.list(x))
    return(NULL)
  unlist(lapply(x, find_df), FALSE)
}

data <- find_df(mylist)

# # add name column to list
f <- function (data, name){
  data$name <- name
  data
}

data <- Map(f, data, names(data))

# convert list to dataframe
data.commu <- as.data.frame(do.call(rbind, data))

# remove ID
data.commu <- data.commu%>%
  filter(ID!="37506237_D_2022-02-01")

data.commu$name[which(data.commu$name=="2D507235_G_2022-02-07_2")] <- "2D507235_G_2022-02-07_1"

data.commu.ls <- split(data.commu, data.commu$name)

lapply(split(data.commu, data.commu$tag_local_identifier), function(x){
 ggplot(aes(x=location_long, y=location_lat, col=ID, group = date), data=x)+
  geom_point()#+
  #theme(legend.position = "none") 
})

# select only outbound commuting
data.commu.out <- data.commu[grep("_1", data.commu$name),]
unique(data.commu.out$name)  


# to check that I selected only outbound commutes
x <- split(data.commu.out, data.commu.out$name)

# plot outbound commutes
lapply(split(data.commu.out, data.commu.out$tag_local_identifier), function(x){
 ggplot(aes(x=location_long, y=location_lat, col=name), data=x)+
  geom_point()#+
  #theme(legend.position = "none") 
})

# select only inbound commuting
data.commu.in <- data.commu[grep("_1", invert=TRUE,data.commu$name),]
unique(data.commu.in$name)  

data.commu.in <- data.commu.in %>%
  group_by(name)%>%
  filter(n>=11 & name!="23506B4E_G_2022-02-01_3")

# to check that I selected only inbound commutes
y <- split(data.commu.in, data.commu.in$name)

# plot inbound commutes
lapply(split(data.commu.in, data.commu.in$tag_local_identifier), function(x){
 ggplot(aes(x=location_long, y=location_lat, col=ID), data=x)+
  geom_point()#+
  #theme(legend.position = "none") 
})
#remove 37506237_D_2022-02-01

unique(data.commu.out$date)
class(data.commu.out$date)

table(data.commu.out$tag_local_identifier, data.commu.out$date)
table(data.commu.out$tag_local_identifier, data.commu.out$sex)

hist(data.commu.out$ground_speed)
```

### Plot outbound commutes to use

```{r, message=FALSE, warning=FALSE}
# pdf(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/analysis/results/dtw/idDay.pdf")
# lapply(data.commu.out, function(x){
#   ggplot(aes(x=coords.x1.1, y=coords.x2.1, color=as.factor(date)), data=x)+
#     geom_point()+
#     ggtitle(x$tag_local_identifier)
# })
# while (!is.null(dev.list()))  dev.off()

```


## Calculating path similarity for each individual

```{r, echo=FALSE}
# split data by day
feb_id <- split(data.commu.out, f = data.commu.out$tag_local_identifier)

# compute DTW distances within each day between different individuals
results_list <- lapply(names(feb_id), function(id) {
  id_data <- feb_id[[id]]
  
  # split data by day within the individual
  indiv_splits <- split(id_data, id_data$date)
  
  # create all pairwise combinations of individuals for the day
  pairs <- expand.grid(names(indiv_splits), names(indiv_splits))
  pairs <- subset(pairs, Var1 != Var2)
  
# compute DTW distance for each pair of individuals
  dtw_distances <- mapply(function(ind1, ind2) {
    dtw_result <- dtw(indiv_splits[[ind1]][, c("coords.x1", "coords.x2")],
                      indiv_splits[[ind2]][, c("coords.x1", "coords.x2")],
                      keep.internals = TRUE)
    dtw_result$distance  # extracting the DTW distance
  }, pairs$Var1, pairs$Var2, SIMPLIFY = FALSE)
  
  # combine the results with pair information and include the date
  data.frame(id= unique(id_data$tag_local_identifier), date1=pairs$Var1, date2=pairs$Var2, Distance = unlist(dtw_distances))
})

names(results_list) <- names(feb_id)

results_df <- do.call(rbind,results_list)

# add columns of groups id and categorical variable of same group= yes/no
dist_id <- results_df %>%
  mutate(group= str_sub(id, 10,10))

dist_id <- dist_id[!duplicated(dist_id$Distance),]

dist_id_ls <- split(dist_id, dist_id$id) 

# cluster by individual
fitall_ind<- lapply(dist_id_ls, function(x){
  hclust(as.dist(xtabs(Distance ~ date1 + date2, data = x)), method = "ward.D2") #as.dist(xtabs(Distance ~ date1 + date2, data = id))
}) # there is something wrong here

names(fitall_ind) <- names(dist_id_ls)

#remove individuals with only two days of tracking
fitall_ind <- fitall_ind[names(fitall_ind) %in% c("53506935_G", "22507B4E_D", "38500437_D", "38506D37_D")==FALSE]

# plot cluster by id in each day
pdf(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/clustering_byiddate_2022.pdf")
lapply(1:length(fitall_ind), function(i){
  id <- names(fitall_ind[i]) 
  fviz_dend(fitall_ind[[i]], k = 2, color_labels_by_k = FALSE, rect = TRUE, main = id)
})
while (!is.null(dev.list()))  dev.off()

pdf(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/tracksidday.pdf")
lapply(1:length(feb_id), function(i){
  id <- names(feb_id[i]) 
  ggplot(feb_id[[id]], aes(x=location_long, y=location_lat, color=as.factor(date), group =as.factor(date)))+
    geom_path()+
    ggtitle(id)
})
while (!is.null(dev.list()))  dev.off()

#plot one individual
ggplot(aes(x=location_long, y=location_lat, group = date, color=as.factor(date)), data=feb_id[["2C507735_G"]])+
geom_path()+
  facet_wrap(.~date)
```

## Plot some tracks per individual of group G to see how the similar are in their commutes

```{r}
#look that there is something going on here
outplots <- data.commu.out %>%
  filter( tag_local_identifier== "2C500235_D"| tag_local_identifier== "2C507735_G" |  tag_local_identifier== "37507537_D"| tag_local_identifier== "39506037_G") %>% #tag_local_identifier== "2D507235_G" | |tag_local_identifier=="2E500235_G"
  ggplot(aes(x=location_long, y=location_lat, col=tag_local_identifier, group=ID))+
  scale_color_manual(values=c( "#1F78B4",  "#CAB2D6", "#33A02C", "#B15928"))+  
  geom_path()+
  xlab("longitude")+
  ylab("latitude")+
  theme_classic()+
  theme(legend.position = "none")+
  theme(axis.title = element_text(size=20), axis.title.x = element_text(size = 20), axis.title.y=element_text(size=20), axis.text.y = element_text(size=14), axis.text.x = element_text(size=14))
ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/tracks_out.pdf", width = 5, height = 5)



inplots <- data.commu.in %>%
  filter( tag_local_identifier== "2C500235_D"| tag_local_identifier== "2C507735_G" | tag_local_identifier== "37507537_D"| tag_local_identifier== "39506037_G") %>%#tag_local_identifier== "2D507235_G" | |tag_local_identifier=="2E500235_G"
  ggplot(aes(x=location_long, y=location_lat, col=tag_local_identifier, group=ID))+
  scale_color_manual(values=c( "#1F78B4",  "#CAB2D6", "#33A02C", "#B15928"))+   
  geom_path()+
  xlab("longitude")+
  ylab("latitude")+
  theme_classic()+
  theme(legend.position = "none")+
  theme(axis.title = element_text(size=20), axis.title.x = element_text(size = 20), axis.title.y=element_text(size=20), axis.text.y = element_text(size=14), axis.text.x = element_text(size=14))
ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/tracks_in.pdf", width = 5, height = 5)

```

## Clustering the outbound commutes by ID day for all individuals in February 2022

Here we use dtw to calculate distance dissimilarities between commuting track by individual day. We calculate all the pairwise comparison between individual in the different days.
Dynamic Time Warping finds the optimal alignment between the two time series by computing the minimal cumulative distance path through this matrix. The resulting alignment stretches or compresses the sequences to match each other as closely as possible, minimizing the total distance (or "warping cost").


To qualitatively described the the commuting paths by individual, I used a clustering method from the hclust function. The clustering method was ward.D2 which calculates minimum variance. It aims to find compact and spherical clusters. This cluster seems to capture better the similarity within individuals compare to the complete linkage which find similar clusters. "ward.D2" implements that criterion (Murtagh and Legendre 2014). With ward.D2, the dissimilarities are squared before cluster updating, compare to ward.D.

```{r}
# split data by day
feb_out <- data.commu.out%>%
  filter(date>="2022-01-31" & date<="2022-02-10")

feb_out_idday <- split(feb_out, f = feb_out$ID)

# function to compute DTW distances between two dataframes
compute_dtw_between_dataframes <- function(df1, df2) {
  dtw_result <- dtw(df1[, c("coords.x2", "coords.x1")],
                    df2[, c("coords.x2", "coords.x1")],
                    keep.internals = TRUE, step.pattern = symmetric2, open.end=TRUE)
  dtw_result$distance  # Extracting the DTW distance
}

# create all pairwise combinations of dataframes
pairs <- expand.grid(names(feb_out_idday), names(feb_out_idday))
pairs <- subset(pairs, Var1 != Var2)

set.seed(1)
# compute DTW distances for each pair of dataframes
dtw_distances <- mapply(function(df1_name, df2_name) {
  df1 <- feb_out_idday[[df1_name]]
  df2 <- feb_out_idday[[df2_name]]
  compute_dtw_between_dataframes(df1, df2)
}, pairs$Var1, pairs$Var2, SIMPLIFY = FALSE)



# combine the results with pair information
all_results <- data.frame(Pair1 = pairs$Var1, Pair2 = pairs$Var2, Distance = unlist(dtw_distances))

# save results
# save(all_results, file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/analysis/data/dtw_feb_bydate.RData")

# add columns of groups id and categorical variable of same group= yes/no
all_results <- all_results %>%
  mutate(id1= str_sub(Pair1, 1,8), id2= str_sub(Pair2, 1,8)) %>%
  mutate(sameid=ifelse(id1 == id2, "yes", "no"), dyad=paste(id1,id2, sep="_"), group1= str_sub(Pair1, 10,10), group2= str_sub(Pair2, 10,10), samegroup=ifelse(group1 == group2, "yes", "no"), group=paste(group1,group2, sep="_"), date1=str_sub(Pair1, 12,21), date2=str_sub(Pair2, 12,21), samedate=ifelse(date1 == date2, "yes", "no"), direc="outbound", season="dry", cave="lagruta")

# change labels of groups
all_results$group[which(all_results$group=="D_G")] <- "G_D"


# Remove duplicates based on 'Distance' column
all_results <- all_results[!duplicated(all_results$Distance), ]

all_results_samedate <- all_results%>%
  filter(sameid=="yes")

#remove individulas with only two days of tracking for the clustering representation
all_results_sub <- all_results %>%
  filter(id1!="53506935" & id1!= "22507B4E" & id1!= "38500437" & id1!="38506D37" & id2!="53506935" & id2!= "22507B4E" & id2!= "38500437" & id2!="38506D37")

all_results_sub$Pair1 <- droplevels(all_results_sub$Pair1)
all_results_sub$Pair2 <- droplevels(all_results_sub$Pair2)

# quickly plot a cluster for all to visualize
set.seed(1)
fit_idday <- hclust(as.dist(xtabs(Distance ~ Pair1 + Pair2, data = all_results_sub)), method = "ward.D2")

# Plot the dendrogram
fviz_dend(fit_idday, k = 15,  color_labels_by_k = TRUE, rect = TRUE,show_labels = TRUE)

ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/cluster_allfeb_idday_ward2_16092024_.pdf", height=20, width =25)
# 
# set.seed(1)
# fit_idday <- hclust(as.dist(xtabs(Distance ~ Pair1 + Pair2, data = all_results)), method = "complete")
# fviz_dend(fit_idday, k=15, color_labels_by_k = TRUE, rect = TRUE,show_labels = TRUE)
# 
# ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/cluster_allfeb_idday_complete19082024_meters.pdf", height=20, width =25)
```
The previous dendogram color by cluster but it is no able to color by individual day even if it falls in a different cluster. I try to do this here in the next figure, to see better where the individuals fall within all the possible clusters inside the tree.

```{r}
library(factoextra)
library(dendextend)
library(viridis) 
library(RColorBrewer)

# Example dendrogram data and clustering (you can replace this with your data)
dend <- as.dendrogram(fit_idday) 

# Define a custom color palette or color vector for bars
color_palette <- c("#A6CEE3","#1F78B4", "#B2DF8A", "#33A02C", "#FB9A99", "#E31A1C", "#FDBF6F", "#FF7F00", "#CAB2D6", "#6A3D9A", "#FFFF99", "#B15928")

# match colors with your labels (labels2)
labels1 <- str_sub(fit_idday$labels, 1,10)
labels1_unique <- unique(labels1)
label_colors1 <- setNames(color_palette[1:length(labels1_unique)], labels1_unique)
colored_labels1 <- label_colors1[labels1]
# 
# # Create the fviz_dend plot
# set.seed(1)
# fviz_dend(dend, k = 12, color_labels_by_k = FALSE, rect = FALSE, show_labels = TRUE,
#           label_cols = colored_labels)

# Overlay colored bars (this needs to be done using base R functions after plotting fviz_dend)
# Adjust margins

pdf(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/dtw_febout_ward2.pdf", height=20, width =25)
par(mar = c(4,1,1,12))

# Plot dendrogram again
dend %>% dendextend::set("labels_cex", 1.2) %>% plot(horiz=TRUE)

# Add colored bars
colored_bars(cbind(colored_labels1), dend, horiz = TRUE, rowLabels = "id", cex.rowLabels=2)

# Add legend for the colors
legend("topleft", legend = labels1_unique, fill = color_palette[1:length(labels1_unique)], text.font = 2, cex = 3)
dev.off()

```

## CLustering the inbound commutes by ID day in Feburary 2022

```{r}
# split data by day
feb_in <- data.commu.in%>%
  filter(date>="2022-01-31" & date<="2022-02-10")

feb_in_idday <- split(feb_in, f = feb_in$ID)

# create all pairwise combinations of dataframes
pairs_in_feb <- expand.grid(names(feb_in_idday), names(feb_in_idday))
pairs_in_feb <- subset(pairs_in_feb, Var1 != Var2)

set.seed(1)
# compute DTW distances for each pair of dataframes
dtw_dist_in_feb <- mapply(function(df1_name, df2_name) {
  df1 <- feb_in_idday[[df1_name]]
  df2 <- feb_in_idday[[df2_name]]
  compute_dtw_between_dataframes(df1, df2)
}, pairs_in_feb$Var1, pairs_in_feb$Var2, SIMPLIFY = FALSE)

# combine the results with pair information
result_dist_feb_in <- data.frame(Pair1 = pairs_in_feb$Var1, Pair2 = pairs_in_feb$Var2, Distance = unlist(dtw_dist_in_feb))

# save results
# save(all_results, file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/analysis/data/dtw_feb_bydate.RData")

# add columns of groups id and categorical variable of same group= yes/no
result_dist_feb_in <- result_dist_feb_in %>%
  mutate(id1= str_sub(Pair1, 1,8), id2= str_sub(Pair2, 1,8)) %>%
  mutate(sameid=ifelse(id1 == id2, "yes", "no"), dyad=paste(id1,id2, sep="_"), group1= str_sub(Pair1, 10,10), group2= str_sub(Pair2, 10,10), samegroup=ifelse(group1 == group2, "yes", "no"), group=paste(group1,group2, sep="_"), date1=str_sub(Pair1, 12,21), date2=str_sub(Pair2, 12,21), samedate=ifelse(date1 == date2, "yes", "no"),  direc="inbound", season="dry", cave="lagruta")

# change labels of groups
result_dist_feb_in$group[which(result_dist_feb_in$group=="D_G")] <- "G_D"

# Remove duplicates based on 'time_diff' column
result_dist_feb_in <- result_dist_feb_in[!duplicated(result_dist_feb_in$Distance), ]

#remove individulas with only two days of tracking for the clustering representation
result_dist_feb_in_sub <- result_dist_feb_in %>%
  filter(id1!="53506935" & id1!= "22507B4E" & id1!= "38500437" & id1!="38506D37" & id2!="53506935" & id2!= "22507B4E" & id2!= "38500437" & id2!="38506D37")

result_dist_feb_in_sub$Pair1 <- droplevels(result_dist_feb_in_sub$Pair1)
result_dist_feb_in_sub$Pair2 <- droplevels(result_dist_feb_in_sub$Pair2)

# quickly plot a cluster for all to visualize
set.seed(1)
fit_idday_in_feb <- hclust(as.dist(xtabs(Distance ~ Pair1 + Pair2, data = result_dist_feb_in_sub)), method = "ward.D2")

# Plot the dendrogram
fviz_dend(fit_idday_in_feb, k = 15,  color_labels_by_k = TRUE, rect = TRUE,show_labels = TRUE, horiz=TRUE, phylo_layout="layout_with_drl")

ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/cluster_feb_in_ward2_trial.pdf", height=20, width =25)
# 
# set.seed(1)
# fit_idday <- hclust(as.dist(xtabs(Distance ~ Pair1 + Pair2, data = all_results)), method = "complete")
# fviz_dend(fit_idday, k=15, color_labels_by_k = TRUE, rect = TRUE,show_labels = TRUE)
# 
# ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/cluster_allfeb_idday_complete19082024_meters.pdf", height=20, width =25)
```
The previous dendogram color by cluster but it is no able to color by individual day even if it falls in a different cluster. I try to do this here in the next figure, to see better where the individuals fall within all the possible clusters inside the tree.

```{r}
library(factoextra)
library(dendextend)

# Example dendrogram data and clustering (you can replace this with your data)
dend_in_feb <- as.dendrogram(fit_idday_in_feb) 

# Define a custom color palette or color vector for bars
color_palette <- c("#A6CEE3","#1F78B4", "#B2DF8A", "#33A02C", "#FB9A99", "#E31A1C", "#FDBF6F", "#FF7F00", "#CAB2D6", "#6A3D9A", "#FFFF99", "#B15928", "grey")
# match colors with your labels (labels2)
labels2 <- str_sub(fit_idday_in_feb$labels, 1,10)
labels2_unique <- unique(labels2)
label_colors <- setNames(color_palette[1:length(labels2_unique)], labels2_unique)
colored_labels2 <- label_colors[labels2]
# 
# # Create the fviz_dend plot
# set.seed(1)
# fviz_dend(dend, k = 12, color_labels_by_k = FALSE, rect = FALSE, show_labels = TRUE,
#           label_cols = colored_labels)

# Overlay colored bars (this needs to be done using base R functions after plotting fviz_dend)
# Adjust margins

pdf(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/dtw_feb_in_ward2.pdf", height=20, width =20)
par(mar = c(4,1,2,12))

# Plot dendrogram again
dend_in_feb %>% dendextend::set("labels_cex", 1.2) %>% plot(horiz=TRUE)#

# Add colored bars
colored_bars(cbind(colored_labels2), dend_in_feb, horiz = TRUE, rowLabels = "id", cex.rowLabels=2)

# Add legend for the colors
legend("topleft", legend = labels2_unique, fill = color_palette[1:length(labels2_unique)], text.font = 2, cex = 2.5)
dev.off()

pdf(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/dtw_feb_outin.pdf", height=20, width =20)
# Adjust margins and plot the first dendrogram
par(mar = c(4, 1, 1, 12))
dend %>% dendextend::set("labels_cex", 1.2) %>% plot(horiz=TRUE)

# Add colored bars and legend for the first dendrogram
colored_bars(cbind(colored_labels1), dend, horiz = TRUE, rowLabels = "id", cex.rowLabels=2)
legend("topleft", legend = labels1_unique, fill = color_palette[1:length(labels1_unique)], text.font = 2, cex = 3)

# Adjust margins and plot the second dendrogram
#par(mar = c(4, 1, 2, 12))
dend_in_feb %>% dendextend::set("labels_cex", 1.2) %>% plot(horiz=TRUE)#xlim=c(0,max(dist(fit_idday_in_feb$height)))

# Add colored bars and legend for the second dendrogram
colored_bars(cbind(colored_labels2), dend_in_feb, horiz = TRUE, rowLabels = "id", cex.rowLabels=2)
#legend("topleft", legend = labels2_unique, fill = color_palette[1:length(labels2_unique)], text.font = 2, cex = 2.5)
dev.off()

```

### Test if there is more similarities within than between individuals in their paths similarities

```{r}
library(lme4)

# check the distribution of distance
library(fitdistrplus)
fit <- fitdist(all_results$Distance, "weibull")  # Change "weibull" to other distributions
summary(fit)
gofstat(fit)

# diagnostic plots
plot(fit)
cdfcomp(fit)
qqcomp(fit)

# fit multiple distributions
fit_norm <- fitdist(all_results$Distance, "norm")
fit_gamma <- fitdist(all_results$Distance, "gamma")
fit_weib <- fitdist(all_results$Distance, "weibull")

# compare fits visually
# assuming fit_norm has estimated parameters mean (mle) and standard deviation (sd)
plotdist(all_results$Distance, "norm", para = list(mean = fit_norm$estimate["mean"], sd = fit_norm$estimate["sd"]), histo = TRUE, demp = TRUE)
cdfcomp(list(fit_norm, fit_weib), legendtext = c("Normal", "Weibull"))
# weibull is the best distribution

# plot outbound similarity between and within ids
ggplot(aes(x=sameid, y=Distance), data=all_results) +
  geom_jitter(aes(alpha=0.5))+
stat_summary(fun = mean, color = "red") +
  stat_summary(
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "red",
    width = .3
  )

ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/dtw_outbound_feb.pdf")

# plot inbound similarity between and within ids
ggplot(aes(x=sameid, y=Distance), data=result_dist_feb_in) +
  geom_jitter(aes(alpha=0.5))+
stat_summary(fun = mean, color = "red") +
  stat_summary(
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "red",
    width = .3
  )

ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/dtw_inbound_feb.pdf")


# test similarity of outbound commutes in Feburary
all_results$date <- paste(all_results$date1, all_results$date2, sep="_")

m_dtw_out <- lmer(log(Distance)~sameid + (1|dyad) + (1|date), data = all_results)

summary(m_dtw_out)
car::Anova(m_dtw_out)

# plot histogram for distance
hist(log(result_dist_feb_in$Distance))

# test similarity of inbound commutes in Feburary
result_dist_feb_in$date <- paste(result_dist_feb_in$date1, result_dist_feb_in$date2, sep="_")

m_dtw_in <- lmer(log(Distance)~sameid + (1|dyad) + (1|date), data = result_dist_feb_in)

summary(m_dtw_in)
car::Anova(m_dtw_in)
```
Here we see that independent of the outbound and inbound commutes, individuals have more similar paths across the days than path between other individuals in the group.

Now we are interested in seeing if outbound commutes are similar than inbound commutes within the same individuals, to see if there are some particular routes preferred when bats fly to the foraging areas vs. returning to the roost.



### Compare inbound and outbound from the same individual same day in February

```{r}
data.commu.out$direc <- "outbound"
data.commu.in$direc <- "inbound"

febpaths <- rbind(data.commu.out,data.commu.in)

febpaths_id <- split(febpaths, febpaths$tag_local_identifier)

# # compute DTW distances within each day between different individuals
# results_inout <- lapply(names(febpaths_id), function(id) {
#   id_data <- febpaths_id[[id]]
#   
#   # split data by day within the individual
#   indiv_splits <- split(id_data, id_data$direc)
#   
#   # create all pairwise combinations of individuals for the day
#   # pairs <- expand.grid(names(indiv_splits), names(indiv_splits))
#   # pairs <- subset(pairs, Var1 != Var2)
#   
# # compute DTW distance for each pair of individuals
#     dtw_result <- dtw(indiv_splits[[1]][, c("coords.x1", "coords.x2")],
#                       indiv_splits[[2]][, c("coords.x1", "coords.x2")],
#                       keep.internals = TRUE)
#     dist <- dtw_result$distance  # extracting the DTW distance
#   
#   # combine the results with pair information and include the date
#   data.frame(id= unique(id_data$tag_local_identifier), date=unique(id_data$date), Distance = unlist(dist))
# })
# 
# names(results_inout) <- names(febpaths_id)
# 
# results_inout_df <- do.call(rbind, results_inout)
# results_inout_df$season <- "2022_feb"


######## try for all pairwise comparisons


# Compute pairwise DTW distances between outbound and inbound commutes across all date combinations
results_inout <- lapply(names(febpaths_id), function(id) {
  id_data <- febpaths_id[[id]]
  
  # Split data by direction (outbound and inbound)
  direction_splits <- split(id_data, id_data$direc)
  
  # Check if both outbound and inbound data exist
  if (!("outbound" %in% names(direction_splits)) || !("inbound" %in% names(direction_splits))) {
    return(NULL) # Skip if one direction is missing
  }
  
  outbound_data <- direction_splits$outbound
  inbound_data <- direction_splits$inbound
  
  # Create all combinations of dates between outbound and inbound commutes
  date_combinations <- expand.grid(outbound_date = unique(outbound_data$date),
                                    inbound_date = unique(inbound_data$date),
                                    stringsAsFactors = FALSE)
  
  # Calculate DTW distance for each combination
  distances <- apply(date_combinations, 1, function(row) {
    out_date <- row["outbound_date"]
    in_date <- row["inbound_date"]
    
    # Filter data for the specific dates
    out_coords <- outbound_data[outbound_data$date == out_date, c("coords.x1", "coords.x2")]
    in_coords <- inbound_data[inbound_data$date == in_date, c("coords.x1", "coords.x2")]
    
    if (nrow(out_coords) > 0 && nrow(in_coords) > 0) {
      # Compute DTW distance
      dtw_result <- dtw(out_coords, in_coords, keep.internals = TRUE)
      return(dtw_result$distance)
    } else {
      return(NA) # Handle cases where no data is available
    }
  })
  
  # Combine the results
  results <- data.frame(
    id = id,
    outbound_date = date_combinations$outbound_date,
    inbound_date = date_combinations$inbound_date,
    Distance = distances
  )
  
  return(results)
})

# Combine results for all individuals
results_inout_df <- do.call(rbind, results_inout)

# View final results
head(results_inout_df)

results_inout_df$season <- "2022_feb"

# plot boxplot of similarities
ggplot(aes(x=id, y=Distance), data=results_inout_df) +
  geom_boxplot()+
  geom_point()+
stat_summary(fun = mean, color = "red") +
  stat_summary(
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "red",
    width = .3
  )+
  ylab("distance similarity")+
  theme_classic()+
  theme(axis.title = element_text(face = 'bold', size=20), axis.title.x = element_text(size = 20), axis.title.y=element_text(size=20), axis.text.y = element_text(size=14), axis.text.x = element_text(size=14))
ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/inout_sim_byid.pdf")

```


## Males in aj cave
```{r}
march_males <- bats_commu%>%
  filter(date>="2022-03-01" & date<="2022-03-30")

splitmarch_males <- split(march_males, f=march_males$ID)

Commu_march_commales <- splitmarch_males[names(splitmarch_males) %in% c("PH_TS_011_2022-03-08","PH_TS_011_2022-03-12", "PH_TS_011_2022-03-15", "PH_TS_011_2022-03-17", "PH_TS_014_2022-03-07", "PH_TS_014_2022-03-08", "PH_TS_014_2022-03-09", "PH_TS_014_2022-03-10", "PH_TS_016_2022-03-09","PH_TS_016_2022-03-10", "PH_TS_080_2022-03-09", "PH_TS_080_2022-03-10", "PH_TS_080_2022-03-11", "PH_TS_080_2022-03-14", "PH_TS_080_2022-03-15", "PH_TS_080_2022-03-16", "PH_TS_080_2022-03-17", "PH_TS_080_2022-03-18", "PH_TS_080_2022-03-19", "PH_TS_080_2022-03-20", "PH_TS_080_2022-03-21", "PH_TS_100_2022-03-07", "PH_TS_100_2022-03-10", "PH_TS_100_2022-03-11", "PH_TS_100_2022-03-16", "PH_TS_100_2022-03-17", "PH_TS_100_2022-03-18", "PH_TS_112_2022-03-08", "PH_TS_112_2022-03-11", "PH_TS_120_2022-03-08") == TRUE]#remove "PH_TS_120_2022-03-08", "PH_TS_120_2022-03-11"

Commu_march_malesc <- as.data.frame(do.call(rbind, Commu_march_commales))

#plot time vs id to see gaps
ggplot(aes(x=location_long, y=location_lat, color=as.factor(date)), data=Commu_march_malesc)+
    geom_point()+
    facet_wrap(.~date)


# ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/paths_males_ajcave.pdf", height=20, width =25)


# for loop to divide each day of the animal in commuting outbound and commuting inbound using around 30 minutes as a threshold value
mylist2 <- list() 
for (i in 1:length(Commu_march_commales)){
  x <- which(Commu_march_commales[[i]]$diff>=1215)
  mylist2[[i]] <- split(Commu_march_commales[[i]], cumsum(1:nrow(Commu_march_commales[[i]])%in%x))
  indv <- unique(names(Commu_march_commales[i]))
  names(mylist2[[i]]) <- paste(indv, seq_along(mylist2[[i]]), sep = "_")
}

# convert nested list to a data frame
data_march <- find_df(mylist2)

# add names to list of the commuting outbound or inbound to the list
data_march <- Map(f, data_march, names(data_march))

# convert list to data frame
data.march <- as.data.frame(do.call(rbind, data_march))

data.march <- data.march %>%
  filter(ID!="PH_TS_011_2022-03-15", ID!="PH_TS_011_2022-03-17", ID!="PH_TS_014_2022-03-08", ID!="PH_TS_014_2022-03-07", ID!= "PH_TS_079_2022-03-16", ID!= "PH_TS_100_2022-03-17", ID!= "PH_TS_100_2022-03-18", ID!= "PH_TS_080_2022-03-09")

# remove ID
data.march <- data.march%>%
  filter(name!="PH_TS_016_2022-03-10_1")

data.march$name[which(data.march$name=="PH_TS_016_2022-03-10_2")] <- "PH_TS_016_2022-03-10_1"
data.march$name[which(data.march$name=="PH_TS_016_2022-03-10_3")] <- "PH_TS_016_2022-03-10_2"

# select only outbound commuting
commu.march.out <- data.march[grep("_1\\b", data.march$name),]#exact match to avoid selecting the inbound commuting tracks
commu.march.out <- commu.march.out%>% 
  filter(ID!="PH_TS_011_2022-03-15", ID!="PH_TS_011_2022-03-17", ID!= "PH_TS_079_2022-03-16", ID!= "PH_TS_100_2022-03-17", ID!= "PH_TS_100_2022-03-18",ID!="PH_TS_120_2022-03-11")

commu.march.in <- data.march[grep("_2\\b", data.march$name),]
commu.march.in <- commu.march.in %>%
  filter( ID!= "PH_TS_079_2022-03-16", ID!= "PH_TS_080_2022-03-09")

# plot outbound commutes
ggplot(commu.march.out, aes(x=location_long, y=location_lat, group=ID, color=ID))+
  geom_path()#+
  #facet_wrap(.~ID)

# plot inbound commutes
ggplot(commu.march.in, aes(x=location_long, y=location_lat, group = ID, color=ID))+
  geom_path()#+
 # facet_wrap(.~ID)
```

## DTW similarity of outbound commutes for males in march 

```{r}
# split dataframe by ID
commu.march.out.ls <- split(commu.march.out, commu.march.out$ID)

# Create all pairwise combinations of dataframes
pairs_out_march <- expand.grid(names(commu.march.out.ls), names(commu.march.out.ls))
pairs_out_march <- subset(pairs_out_march, Var1 != Var2)

set.seed(1)

# compute DTW distances for each pair of dataframes
dtw_distances_marchout <- mapply(function(df1_name, df2_name) {
  df1 <- commu.march.out.ls[[df1_name]]
  df2 <- commu.march.out.ls[[df2_name]]
  compute_dtw_between_dataframes(df1, df2)
}, pairs_out_march$Var1, pairs_out_march$Var2, SIMPLIFY = FALSE)

# combine the results with pair information
results_march_out <- data.frame(Pair1 = pairs_out_march$Var1, Pair2 = pairs_out_march$Var2, Distance = unlist(dtw_distances_marchout))

# save results
# save(all_results, file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/analysis/data/dtw_feb_bydate.RData")

# add columns of groups id and categorical variable of same group= yes/no
results_march_out<- results_march_out %>%
  mutate(id1= str_sub(Pair1, 1,9), id2= str_sub(Pair2, 1,9)) %>%
  mutate(sameid=ifelse(id1 == id2, "yes", "no"), dyad=paste(id1,id2, sep="_"), date1=str_sub(Pair1, 11,20), date2=str_sub(Pair2, 11,20), samedate=ifelse(date1 == date2, "yes", "no"), direc="outbound", , season="dry", cave="ajcave")

# Remove duplicates based on 'Distance' column
results_march_out <- results_march_out[!duplicated(results_march_out$Distance), ]

march_out_samedate <- results_march_out%>%
  filter(samedate=="yes")

# quickly plot a cluster for all to visualize
set.seed(1)
fit_idday_marchout <- hclust(as.dist(xtabs(Distance ~ Pair1 + Pair2, data = results_march_out)), method = "ward.D2")

# Plot the dendrogram
fviz_dend(fit_idday_marchout, k = 10,  color_labels_by_k = TRUE, rect = TRUE,show_labels = TRUE, cex = 2)

ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/cluster_march_out_ward2.pdf", height=35, width =25)

#still test this
# boxplot of similarities in outbound commutes in march
ggplot(aes(x=sameid, y=Distance), data=results_march_out) +
  geom_jitter(aes(alpha=0.5))+
stat_summary(fun = mean, color = "red") +
  stat_summary(
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "red",
    width = .3
  )
```

```{r}
# Example dendrogram data and clustering (you can replace this with your data)
dend_marchout <- as.dendrogram(fit_idday_marchout) 

# Define a custom color palette or color vector for bars
color_palette2 <- c("#A6CEE3","#1F78B4", "#B2DF8A", "#33A02C", "#FB9A99", "#E31A1C", "#FDBF6F", "#FF7F00", "#CAB2D6")

# match colors with your labels (labels2)
labels_marchout <- str_sub(fit_idday_marchout$labels, end=-12)
label_uni_marchout <- unique(labels_marchout)
label_colors_marchout <- setNames(color_palette2[1:length(label_uni_marchout)], label_uni_marchout)
colored_labels_march <- label_colors_marchout[labels_marchout]


pdf(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/dtw_marchout_ward2.pdf", height=20, width =25)
par(mar = c(4,1,1,12))

# Plot dendrogram again
dend_marchout %>% dendextend::set("labels_cex", 1.2) %>% plot(horiz=TRUE)

# Add colored bars
colored_bars(cbind(colored_labels_march), dend_marchout, horiz = TRUE, rowLabels = "id")

# Add legend for the colors
legend("topleft", legend = label_uni_marchout, fill = color_palette2[1:length(label_uni_marchout)], text.font = 2, cex = 3)
dev.off()

```
## DTW similarity of inbound commutes for males in march 

```{r}
# split dataframe by ID
commu.march.in.ls <- split(commu.march.in, commu.march.in$ID)

# Create all pairwise combinations of dataframes
pairs_in_march <- expand.grid(names(commu.march.in.ls), names(commu.march.in.ls))
pairs_in_march <- subset(pairs_in_march, Var1 != Var2)

set.seed(1)

# compute DTW distances for each pair of dataframes
dtw_distances_marchin <- mapply(function(df1_name, df2_name) {
  df1 <- commu.march.in.ls[[df1_name]]
  df2 <- commu.march.in.ls[[df2_name]]
  compute_dtw_between_dataframes(df1, df2)
}, pairs_in_march$Var1, pairs_in_march$Var2, SIMPLIFY = FALSE)

# combine the results with pair information
results_march_in <- data.frame(Pair1 = pairs_in_march$Var1, Pair2 = pairs_in_march$Var2, Distance = unlist(dtw_distances_marchin))

# save results
# save(all_results, file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/analysis/data/dtw_feb_bydate.RData")

# add columns of groups id and categorical variable of same group= yes/no
results_march_in<- results_march_in %>%
  mutate(id1= str_sub(Pair1, 1,9), id2= str_sub(Pair2, 1,9)) %>%
  mutate(sameid=ifelse(id1 == id2, "yes", "no"), dyad=paste(id1,id2, sep="_"), date1=str_sub(Pair1, 11,20), date2=str_sub(Pair2, 11,20), samedate=ifelse(date1 == date2, "yes", "no"), direc="inbound", , season="dry", cave="ajcave")

# Remove duplicates based on 'time_diff' column
results_march_in <- results_march_in[!duplicated(results_march_in$Distance), ]


march_in_samedate <- results_march_in%>%
  filter(samedate=="yes")

# quickly plot a cluster for all to visualize
set.seed(1)
fit_idday_marchin <- hclust(as.dist(xtabs(Distance ~ Pair1 + Pair2, data = results_march_in)), method = "ward.D2")

# Plot the dendrogram
fviz_dend(fit_idday_marchin, k = 10,  color_labels_by_k = TRUE, rect = TRUE,show_labels = TRUE, cex = 2)

ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/cluster_march_in_ward2.pdf", height=35, width =25)

#still test this
# boxplot of similarities in outbound commutes in march
ggplot(aes(x=sameid, y=Distance), data=results_march_in) +
  geom_jitter(aes(alpha=0.5))+
stat_summary(fun = mean, color = "red") +
  stat_summary(
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "red",
    width = .3
  )
```

```{r}
# Example dendrogram data and clustering (you can replace this with your data)
dend_marchin <- as.dendrogram(fit_idday_marchin) 

# Define a custom color palette or color vector for bars
color_palette3 <- c("#A6CEE3","#1F78B4", "#B2DF8A", "#33A02C", "#FB9A99", "#E31A1C", "#FDBF6F", "#FF7F00", "#CAB2D6")

# match colors with your labels (labels2)
labels_marchin <- str_sub(fit_idday_marchin$labels, end=-12)
label_uni_marchin <- unique(labels_marchin)
label_colors_marchin <- setNames(color_palette2[1:length(label_uni_marchin)], label_uni_marchin)
colored_labels_marchin <- label_colors_marchin[labels_marchin]


pdf(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/dtw_marchin_ward2.pdf", height=20, width =25)
par(mar = c(4,1,1,12))

# Plot dendrogram again
dend_marchin %>% set("labels_cex", 1.2) %>% plot(horiz=TRUE)

# Add colored bars
colored_bars(cbind(colored_labels_marchin), dend_marchin, horiz = TRUE, rowLabels = "id")

# Add legend for the colors
legend("topleft", legend = label_uni_marchin, fill = color_palette2[1:length(label_uni_marchin)], text.font = 2, cex = 3)
dev.off()
```


### Compaare inbound and outbound from the same individual same day in March

```{r}
commu.march.out$direc <- "outbound"
commu.march.in$direc <- "inbound"

marchpaths <- rbind(commu.march.out,commu.march.in)

marchpaths_id <- split(marchpaths, marchpaths$tag_local_identifier)

# Compute pairwise DTW distances between outbound and inbound commutes across all date combinations
results_inout_march <- lapply(names(marchpaths_id), function(id) {
  id_data <- marchpaths_id[[id]]
  
  # Split data by direction (outbound and inbound)
  direction_splits <- split(id_data, id_data$direc)
  
  # Check if both outbound and inbound data exist
  if (!("outbound" %in% names(direction_splits)) || !("inbound" %in% names(direction_splits))) {
    return(NULL) # Skip if one direction is missing
  }
  
  outbound_data <- direction_splits$outbound
  inbound_data <- direction_splits$inbound
  
  # Create all combinations of dates between outbound and inbound commutes
  date_combinations <- expand.grid(outbound_date = unique(outbound_data$date),
                                    inbound_date = unique(inbound_data$date),
                                    stringsAsFactors = FALSE)
  
  # Calculate DTW distance for each combination
  distances <- apply(date_combinations, 1, function(row) {
    out_date <- row["outbound_date"]
    in_date <- row["inbound_date"]
    
    # Filter data for the specific dates
    out_coords <- outbound_data[outbound_data$date == out_date, c("coords.x1", "coords.x2")]
    in_coords <- inbound_data[inbound_data$date == in_date, c("coords.x1", "coords.x2")]
    
    if (nrow(out_coords) > 0 && nrow(in_coords) > 0) {
      # Compute DTW distance
      dtw_result <- dtw(out_coords, in_coords, keep.internals = TRUE)
      return(dtw_result$distance)
    } else {
      return(NA) # Handle cases where no data is available
    }
  })
  
  # Combine the results
  results <- data.frame(
    id = id,
    outbound_date = date_combinations$outbound_date,
    inbound_date = date_combinations$inbound_date,
    Distance = distances
  )
  
  return(results)
})

# Combine results for all individuals
results_inout_march_df <- do.call(rbind, results_inout_march)

# View final results
head(results_inout_march_df)

# compute DTW distances within each day between different individuals
# results_inout_march <- lapply(names(marchpaths_id), function(id) {
#   id_data <- marchpaths_id[[id]]
#   
#   # split data by day within the individual
#   indiv_splits <- split(id_data, id_data$direc)
#   
#   # create all pairwise combinations of individuals for the day
#   # pairs <- expand.grid(names(indiv_splits), names(indiv_splits))
#   # pairs <- subset(pairs, Var1 != Var2)
#   
# # compute DTW distance for each pair of individuals
#     dtw_result <- dtw(indiv_splits[[1]][, c("coords.x1", "coords.x2")],
#                       indiv_splits[[2]][, c("coords.x1", "coords.x2")],
#                       keep.internals = TRUE)
#     dist <- dtw_result$distance  # extracting the DTW distance
#   
#   # combine the results with pair information and include the date
#   data.frame(id= unique(id_data$tag_local_identifier), date=unique(id_data$date), Distance = unlist(dist))
# })

# names(results_inout_march) <- names(marchpaths_id)

# results_inout_march_df <- do.call(rbind, results_inout_march)
results_inout_march_df$season <- "2022_march"

# plot boxplot of similarities
ggplot(aes(x=id, y=Distance), data=results_inout_march_df) +
  geom_boxplot()+
  geom_point()+
stat_summary(fun = mean, color = "red") +
  stat_summary(
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "red",
    width = .3
  )+
  ylab("distance similarity")+
  theme_classic()
ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/inout_sim_byid_march.pdf", height=10, width =15)

#plot commuting paths
ggplot(marchpaths[marchpaths$tag_local_identifier=="PH_TS_016",], aes(x=location_long, y=location_lat, color=tag_local_identifier, linetype =direc, sgroup=ID))+
  geom_path()+
  theme_classic()

```



## Filter wet season commuting paths

Here we filtered wet season commuting paths which are complete in other to calculate DTW path similarities.

```{r}
wet_bats <- bocas%>%
  filter(behav=="commuting", n>=5)%>%
  filter(date>="2022-03-30")

# add time difference to split commuting tracks by the ID_day
wet_bats <- wet_bats %>%
  group_by(ID_batday) %>%
  mutate(diff = timestamps - lag(timestamps))
wet_bats$diff <- as.numeric(wet_bats$diff)*60
wet_bats$diff[is.na(wet_bats$diff)] <- 0

# split by ID
splitwet <- split(wet_bats, f=wet_bats$ID)

#complete tracks
commu_wet_complete <- splitwet[names(splitwet) %in% c("PHYL11_2023-08-15", "PHYL11_2023-08-16", "PHYL11_2023-08-17", "PHYL11_2023-08-18", "PHYL16_2023-08-14", "PHYL16_2023-08-15",  "PHYL21_2023-08-18", "PHYL25_2023-08-14", "PHYL25_2023-08-15", "PHYL25_2023-08-16", "PHYL25_2023-08-17", "PHYL25_2023-08-18", "PHYL28_2023-08-14", "PHYL28_2023-08-15", "PHYL34_2023-08-13", "PHYL34_2023-08-14",  "PHYL4_2023-08-14", "PHYL7_2023-08-16", "PHYL7_2023-08-17", "PHYL7_2023-08-18","PHYL9_2023-08-14",    "PHYL9_2023-08-18") == TRUE]

# for loop to divide each day of the animal in commuting outbound and commuting inbound using around 30 minutes as a threshold value
mylist3 <- list() 
for (i in 1:length(commu_wet_complete)){
  x <- which(commu_wet_complete[[i]]$diff>=1900)
  mylist3[[i]] <- split(commu_wet_complete[[i]], cumsum(1:nrow(commu_wet_complete[[i]])%in%x))
  indv <- unique(names(commu_wet_complete[i]))
  names(mylist3[[i]]) <- paste(indv, seq_along(mylist3[[i]]), sep = "_")
}

# convert nested list to a data frame
data_wet <- find_df(mylist3)

# add names to list of the commuting outbound or inbound to the list
data_wet <- Map(f, data_wet, names(data_wet))

# convert list to data frame
data.commu.wet <- as.data.frame(do.call(rbind, data_wet))

# filter ID incomplete commutes
data.commu.wet <- data.commu.wet %>%
  filter(ID!="PHYL28_2023-08-14", ID!="PHYL28_2023-08-15", ID!="PHYL34_2023-08-13", ID!="PHYL34_2023-08-14", ID!="PHYL9_2023-08-14", ID!="PHYL9_2023-08-18", ID!="PHYL4_2023-08-14", ID!="PHYL1_2023-08-14", ID!="PHYL25_2023-08-17", ID!="PHYL9_2023-08-18" ,ID!="PHYL7_2023-08-18")

# select only outbound commuting
data.commu.wet.out <- data.commu.wet[grep("_1\\b", data.commu.wet$name),]#exact match to avoid selecting the inbound commuting tracks

ggplot(data.commu.wet.out, aes(x=location_long, y=location_lat, group = ID, color=ID))+
  geom_path()+
  facet_wrap(.~tag_local_identifier)

data.commu.wet.in <- data.commu.wet[grep("_2\\b", data.commu.wet$name),]
data.commu.wet.in <- data.commu.wet.in%>%
  filter()

ggplot(data.commu.wet.in, aes(x=location_long, y=location_lat, group = ID, color=ID))+
  geom_path()+
  facet_wrap(.~tag_local_identifier)

mapview::mapView(data.commu.wet.out, xcol="location_long", ycol="location_lat", zcol="ID", legend=F, crs="EPSG:4326")


mapview::mapView(data.commu.wet.in, xcol="location_long", ycol="location_lat", zcol="ID", legend=F, crs="EPSG:4326")

```

## DTW for outbound commutes paths in the wet season of 2023

```{r}
#split commuting outbound paths
commu.wet.out.ls <- split(data.commu.wet.out, data.commu.wet.out$ID)

# Create all pairwise combinations of dataframes
pairs_wet_out <- expand.grid(names(commu.wet.out.ls), names(commu.wet.out.ls))
pairs_wet_out <- subset(pairs_wet_out, Var1 != Var2)

set.seed(1)
# Compute DTW distances for each pair of dataframes
distances_wet_out <- mapply(function(df1_name, df2_name) {
  df1 <- commu.wet.out.ls[[df1_name]]
  df2 <- commu.wet.out.ls[[df2_name]]
  compute_dtw_between_dataframes(df1, df2)
}, pairs_wet_out$Var1, pairs_wet_out$Var2, SIMPLIFY = FALSE)

# Combine the results with pair information
results_wet_out <- data.frame(Pair1 = pairs_wet_out$Var1, Pair2 = pairs_wet_out$Var2, Distance = unlist(distances_wet_out))

# save results
# save(all_results, file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/analysis/data/dtw_feb_bydate.RData")

# add columns of groups id and categorical variable of same group= yes/no
results_wet_out<- results_wet_out %>%
  mutate(id1= str_sub(Pair1, end=-12), id2= str_sub(Pair2, end=-12)) %>%
  mutate(sameid=ifelse(id1 == id2, "yes", "no"), dyad=paste(id1,id2, sep="_"), date1=str_sub(Pair1, start=-10), date2=str_sub(Pair2, start=-10), samedate=ifelse(date1 == date2, "yes", "no"), direc="outbound", season="wet")

results_wet_out$cave1 <- data.commu.wet.out$cave[match(results_wet_out$id1, data.commu.wet.out$tag_local_identifier,)]
results_wet_out$cave2 <-data.commu.wet.out$cave[match(results_wet_out$id2, data.commu.wet.out$tag_local_identifier,)]

results_wet_out$samecave <- ifelse(results_wet_out$cave1 == results_wet_out$cave2, "yes", "no")

# Remove duplicates based on 'time_diff' column
results_wet_out <- results_wet_out[!duplicated(results_wet_out$Distance), ]

results_wet_out_samecave <- results_wet_out%>%
  filter(samecave=="yes")

# quickly plot a cluster for all to visualize
set.seed(1)
fit_idday_wet_out <- hclust(as.dist(xtabs(Distance ~ Pair1 + Pair2, data = results_wet_out_samecave)), method = "ward.D2")

# Plot the dendrogram
fviz_dend(fit_idday_wet_out, k =5,  color_labels_by_k = TRUE, rect = TRUE,show_labels = TRUE)

ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/cluster_wet_out_ward2.pdf", height=35, width =25)


# plot boxplot of similarities
ggplot(aes(x=sameid, y=Distance), data=results_wet_out_samecave) +
  geom_jitter(aes(alpha=0.5))+
stat_summary(fun = mean, color = "red") +
  stat_summary(
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "red",
    width = .3
  )

```

## DTW for inbound commutes paths in the wet season of 2023

```{r}
#split commuting outbound paths
commu.wet.in.ls <- split(data.commu.wet.in, data.commu.wet.in$ID)

# Create all pairwise combinations of dataframes
pairs_wet_in <- expand.grid(names(commu.wet.in.ls), names(commu.wet.in.ls))
pairs_wet_in <- subset(pairs_wet_in, Var1 != Var2)

set.seed(1)
# Compute DTW distances for each pair of dataframes
distances_wet_in <- mapply(function(df1_name, df2_name) {
  df1 <- commu.wet.in.ls[[df1_name]]
  df2 <- commu.wet.in.ls[[df2_name]]
  compute_dtw_between_dataframes(df1, df2)
}, pairs_wet_in$Var1, pairs_wet_in$Var2, SIMPLIFY = FALSE)

# Combine the results with pair information
results_wet_in <- data.frame(Pair1 = pairs_wet_in$Var1, Pair2 = pairs_wet_in$Var2, Distance = unlist(distances_wet_in))

# save results
# save(all_results, file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/analysis/data/dtw_feb_bydate.RData")

# add columns of groups id and categorical variable of same group= yes/no
results_wet_in<- results_wet_in %>%
  mutate(id1= str_sub(Pair1, end=-12), id2= str_sub(Pair2, end=-12)) %>%
  mutate(sameid=ifelse(id1 == id2, "yes", "no"), dyad=paste(id1,id2, sep="_"), date1=str_sub(Pair1, start=-10), date2=str_sub(Pair2, start=-10), samedate=ifelse(date1 == date2, "yes", "no"), direc="inbound", season="wet")

results_wet_in$cave1 <- data.commu.wet.in$cave[match(results_wet_in$id1, data.commu.wet.in$tag_local_identifier,)]
results_wet_in$cave2 <-data.commu.wet.in$cave[match(results_wet_in$id2, data.commu.wet.in$tag_local_identifier,)]
results_wet_in$cave <- ifelse(results_wet_in$cave1==results_wet_in$cave2, results_wet_in$cave1, "mixed")

results_wet_in$samecave <- ifelse(results_wet_in$cave1 == results_wet_in$cave2, "yes", "no")

# Remove duplicates based on 'Distance' column
results_wet_in <- results_wet_in[!duplicated(results_wet_in$Distance), ]

results_wet_in_samecave <- results_wet_in%>%
  filter(samecave=="yes")

# quickly plot a cluster for all to visualize
set.seed(1)
fit_idday_in_out <- hclust(as.dist(xtabs(Distance ~ Pair1 + Pair2, data = results_wet_in_samecave)), method = "ward.D2")

# Plot the dendrogram
fviz_dend(fit_idday_in_out, k =5,  color_labels_by_k = TRUE, rect = TRUE,show_labels = TRUE)

ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/cluster_wet_in_ward2.pdf", height=35, width =25)


# plot boxplot of similarities
ggplot(aes(x=sameid, y=Distance), data=results_wet_in_samecave) +
  geom_jitter(aes(alpha=0.5))+
stat_summary(fun = mean, color = "red") +
  stat_summary(
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "red",
    width = .3
  )

```

### Compare inbound and outbound from the same individual same day in Wet season 2023

```{r}
data.commu.wet.out$direc <- "outbound"
data.commu.wet.in$direc <- "inbound"

wetpaths <- rbind(data.commu.wet.out,data.commu.wet.in)

wetpaths_id <- split(wetpaths, wetpaths$tag_local_identifier)

# Compute pairwise DTW distances between outbound and inbound commutes across all date combinations
results_inout_wet <- lapply(names(wetpaths_id), function(id) {
  id_data <- wetpaths_id[[id]]
  
  # Split data by direction (outbound and inbound)
  direction_splits <- split(id_data, id_data$direc)
  
  # Check if both outbound and inbound data exist
  if (!("outbound" %in% names(direction_splits)) || !("inbound" %in% names(direction_splits))) {
    return(NULL) # Skip if one direction is missing
  }
  
  outbound_data <- direction_splits$outbound
  inbound_data <- direction_splits$inbound
  
  # Create all combinations of dates between outbound and inbound commutes
  date_combinations <- expand.grid(outbound_date = unique(outbound_data$date),
                                    inbound_date = unique(inbound_data$date),
                                    stringsAsFactors = FALSE)
  
  # Calculate DTW distance for each combination
  distances <- apply(date_combinations, 1, function(row) {
    out_date <- row["outbound_date"]
    in_date <- row["inbound_date"]
    
    # Filter data for the specific dates
    out_coords <- outbound_data[outbound_data$date == out_date, c("coords.x1", "coords.x2")]
    in_coords <- inbound_data[inbound_data$date == in_date, c("coords.x1", "coords.x2")]
    
    if (nrow(out_coords) > 0 && nrow(in_coords) > 0) {
      # Compute DTW distance
      dtw_result <- dtw(out_coords, in_coords, keep.internals = TRUE)
      return(dtw_result$distance)
    } else {
      return(NA) # Handle cases where no data is available
    }
  })
  
  # Combine the results
  results <- data.frame(
    id = id,
    outbound_date = date_combinations$outbound_date,
    inbound_date = date_combinations$inbound_date,
    Distance = distances
  )
  
  return(results)
})

# Combine results for all individuals
results_inout_wet_df <- do.call(rbind, results_inout_wet)

# compute DTW distances within each day between different individuals
# results_inout_wet <- lapply(names(wetpaths_id), function(id) {
#   id_data <- wetpaths_id[[id]]
#   
#   # split data by day within the individual
#   indiv_splits <- split(id_data, id_data$direc)
#   
#   # create all pairwise combinations of individuals for the day
#   # pairs <- expand.grid(names(indiv_splits), names(indiv_splits))
#   # pairs <- subset(pairs, Var1 != Var2)
#   
# # compute DTW distance for each pair of individuals
#     dtw_result <- dtw(indiv_splits[[1]][, c("coords.x1", "coords.x2")],
#                       indiv_splits[[2]][, c("coords.x1", "coords.x2")],
#                       keep.internals = TRUE)
#     dist <- dtw_result$distance  # extracting the DTW distance
#   
#   # combine the results with pair information and include the date
#   data.frame(id= unique(id_data$tag_local_identifier), date=unique(id_data$date), Distance = unlist(dist))
# })

# names(results_inout_wet) <- names(wetpaths_id)

# results_inout_wet_df <- do.call(rbind, results_inout_wet)
results_inout_wet_df$season <- "2023_august"

# plot boxplot of similarities
ggplot(aes(x=id, y=Distance), data=results_inout_wet_df) +
  geom_boxplot()+
  geom_point()+
stat_summary(fun = mean, color = "red") +
  stat_summary(
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "red",
    width = .3
  )+
  ylab("distance similarity")+
  theme_classic()
ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/inout_sim_byid_2023.pdf", height=10, width =15)

ggplot(wetpaths, aes(x=location_long, y=location_lat, color=tag_local_identifier, linetype =direc, sgroup=ID))+
  geom_path()+
  theme_classic() 


```
### Save all outbound and inbound commutes
```{r}
allpaths <- rbind(febpaths,marchpaths, wetpaths)

#look at sex by id
table(allpaths$tag_local_identifier, allpaths$sex)

# write.csv(allpaths,file = "/Users/ccalderon/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/data/allout_inbound_paths.csv")
write.csv(allpaths,file = "/Users/ccalderon/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/data/allpairwise_out_inbound_paths.csv")

allpaths <- read.csv(file = "/Users/ccalderon/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/data/allpairwise_out_inbound_paths.csv")
```

### in and outbound comparision in distance across seasons

```{r}
#install.packages("gghighlight")
library(gghighlight)
inout_all <- rbind(results_inout_df, results_inout_march_df, results_inout_wet_df)

# add sex
inout_all$sex <- bocas$sex[match(inout_all$id, bocas$tag_local_identifier)]
inout_all$cave_year <- bocas$year_cave[match(inout_all$id, bocas$tag_local_identifier)]
inout_all$samedate <- ifelse(inout_all$outbound_date==inout_all$inbound_date, "yes","no")

ids <- c("37507537_D","39506037_G","PH_TS_080", "PH_TS_100", "PHYL16", "PHYL25")


inout_all_mean <- inout_all %>%
  group_by(cave_year)%>%
  dplyr::summarise(mean_dist=mean(Distance), sd_dist=sd(Distance))

#plot distances
outin_id <- inout_all%>%
  filter(samedate=="yes")%>%
 ggplot(aes(x=season, y=Distance)) +
  geom_boxplot()+#"#A6D854", "#4DAF4A"
  geom_jitter(aes(alpha=0.8, shape = sex), position = position_jitterdodge(jitter.width = 0.4))+
   geom_point(aes(shape = sex), data = inout_all %>% filter(samedate=="yes", id == "37507537_D"), color = "#A6D854", position = position_jitter(width = 0.4), size=3) +
    geom_point(aes(shape = sex),data = inout_all %>% filter(samedate=="yes", id == "39506037_G"), color = "#4DAF4A", position = position_jitter(width = 0.4), size=3) +
    geom_point(aes(shape = sex),data = inout_all %>% filter(samedate=="yes", id == "PH_TS_080"), color = "#756BB1", position = position_jitter(width = 0.4), size=3) +
   geom_point(aes(shape = sex),data = inout_all %>% filter(samedate=="yes", id == "PH_TS_100"), color = "#54278F", position = position_jitter(width = 0.4), size=3) +
  geom_point(aes(shape = sex),data = inout_all %>% filter(samedate=="yes", id == "PHYL11"), color = "#E6AB02", position = position_jitter(width = 0.4), size=3) +
  geom_point(aes(shape = sex),data = inout_all %>% filter(samedate=="yes", id == "PHYL25"), color = "#D95F02", , position = position_jitter(width = 0.4), size=3) +
  #"#CAB2D6","#B15928"
  scale_x_discrete(labels = c("dry 2022-1", "dry 2022-2", "wet 2023"))+#"dry 2022-1", "dry 2022-2", "wet 2023"
#facet_wrap(.~ season)+
stat_summary(fun = mean, color = "red") +
  stat_summary(
    fun.min = function(x) mean(x) - sd(x),
    fun.max = function(x) mean(x) + sd(x),
    geom = "errorbar",
    color = "red",
    width = .3
  )+
  #scale_color_manual(values =c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00"))+
  #gghighlight(id== ids, label_key=ids)+
  ylab("DTW distance between \n out / inbound commutes \n per id")+
  xlab("Period")+
  theme_classic()+
  theme(legend.position = "none", axis.title = element_text(size=20), axis.title.x = element_text(size = 20), axis.title.y=element_text(size=20), axis.text.y = element_text(size=14), axis.text.x = element_text(size=14))

outin_id

ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/inout_across_seasons.pdf")
```

### Plot inbound and outbound commute from the same individuals in different nights

```{r}
library(patchwork)

# Define common limits for longitude and latitude
x_limits <- c(-82.50, -82.28)  # Adjust as needed
y_limits <- c(9.37, 9.44)    # Adjust as needed
y_breaks <- seq(9.37, 9.44, by = 0.02)

#display.brewer.pal(n = 8, name = 'Dark2')
p1 <- febpaths %>%
  filter(tag_local_identifier=="39506037_G" | tag_local_identifier=="37507537_D") %>%
  ggplot(aes(x=location_long, y=location_lat, color=tag_local_identifier, linetype =direc, sgroup=ID))+
  scale_color_manual(values =c("#A6D854", "#4DAF4A"))+ #"#E69F00","#56B4E9"
  geom_path(size=1)+
  xlab("Longitude")+
  ylab("Latitude")+
  scale_x_continuous(limits = x_limits) +
  scale_y_continuous(limits = y_limits, breaks = y_breaks) +
  theme_classic()+
  theme(legend.position = "none", axis.title = element_text(size=20), axis.title.x = element_text(size = 20), axis.title.y=element_text(size=20), axis.text.y = element_text(size=14), axis.text.x = element_text(size=14))


p2 <- marchpaths %>%
  filter(tag_local_identifier=="PH_TS_080" | tag_local_identifier=="PH_TS_014") %>%
  ggplot(aes(x=location_long, y=location_lat, color=tag_local_identifier, linetype =direc, sgroup=ID))+
  scale_color_manual(values =c("#756BB1", "#54278F"))+ #"#009E73","#F0E442"
  geom_path(size=1)+
  xlab("Longitude")+
  ylab("Latitude")+
  scale_x_continuous(limits = x_limits) +
  scale_y_continuous(limits = y_limits, breaks = y_breaks) +
  theme_classic()+
  theme(legend.position = "none", axis.title = element_text(size=20), axis.title.x = element_text(size = 20), axis.title.y=element_text(size=20), axis.text.y = element_text(size=14), axis.text.x = element_text(size=14))



p3 <- wetpaths %>%
  filter(tag_local_identifier=="PHYL25" | tag_local_identifier=="PHYL7") %>%
  ggplot(aes(x=location_long, y=location_lat, color=tag_local_identifier, linetype =direc, sgroup=ID))+
  scale_color_manual(values =c("#E6AB02", "#D95F02"))+ #"#0072B2","#D55E00"
  geom_path(size=1)+
  xlab("Longitude")+
  ylab("Latitude")+
  scale_x_continuous(limits = x_limits) +
  scale_y_continuous(limits = y_limits, breaks = y_breaks) +
  theme_classic()+
  theme(legend.position = "none", axis.title = element_text(size=20), axis.title.x = element_text(size = 20), axis.title.y=element_text(size=20), axis.text.y = element_text(size=14), axis.text.x = element_text(size=14))


tracks <- p1/p2/p3/ plot_layout(axes = "collect")

tracks
```


# Does the difference in outbound and inbound commutes is more different than to the distance of both outbound and inbound commutes?
```{r}
library(lme4)
distsim_all <- all_results %>%
  bind_rows(result_dist_feb_in, results_march_out, results_march_in,results_wet_out,results_wet_in) %>%
  dplyr::select("Pair1", "Pair2", "Distance" , "id1", "id2", "sameid", "dyad", "date1", "date2", "samedate", "direc",  "season", "cave")

distsim_all$year_cave <- "dry 2022-1"

distsim_all$year_cave[which(distsim_all$season=="dry" & distsim_all$cave=="ajcave")] <- "dry 2022-2"

distsim_all$year_cave[which(distsim_all$season=="wet")] <- "wet 2023"

write.csv(distsim_all, file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/data/distsimtable_commutes.csv")
#distsim_all <- read.csv(file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/data/distsimtable_commutes.csv")

distsim_all$sameid <- factor(distsim_all$sameid, levels=c("yes","no"))

#inout_all$day_year <- yday(inout_all$date)
inout_all$direc <- "both"
inout_all$cave_year[inout_all$cave_year=="2022_lagruta_Feb"] <- "dry 2022-1"
inout_all$cave_year[inout_all$cave_year=="2022_ajcave"] <- "dry 2022-2"
inout_all$cave_year[inout_all$cave_year=="2023_ajcave"] <- "wet 2023"
colnames(inout_all)[7] <- "year_cave"

#filter only same id data
distsim_all_sameid <- distsim_all%>%
  filter(sameid=="yes")

distsim_all_sameid$date <- paste(distsim_all_sameid$date1, distsim_all_sameid$date2, sep="_")
distsim_all_sameid$id <- str_sub(distsim_all_sameid$Pair1, end = -12)
distsim_all_sameid$sex <- inout_all$sex[match(distsim_all_sameid$id,inout_all$id)]


# merge distance from same id to the outbound and inbound distance difference by individual by night
inout_all <- rbind(inout_all[,c("id", "Distance", "direc", "year_cave", "sex")], distsim_all_sameid[,c("id", "Distance", "direc", "year_cave", "sex")])


m_diff_inout <- lmer(log(Distance)~year_cave*direc +(1|id), data=inout_all)

summary(m_diff_inout)
car::Anova(m_diff_inout)
library(MuMIn)
r2 <- r.squaredGLMM(m_diff_inout)
print(r2)

plot(m_diff_inout)

acf(residuals(m_diff_inout))

# Marginal means for the interaction of sameid and season
library(emmeans)
emm_interaction <- emmeans(m_diff_inout, ~ direc*year_cave)

# Pairwise comparisons for the interaction
pairwise_interaction <- pairs(emm_interaction, adjust = "tukey")
pairwise_interaction

mean_dist_inout <- inout_all%>%
  group_by(year_cave, direc)%>%
  dplyr::summarise(meandist=mean(Distance, na.rm = TRUE), sddist=sd(Distance, na.rm = TRUE))
```


## Plot similarity within and between individuals of the outbound and inbound commutes

```{r}
library(ggh4x)# for dinamic color in facewraps

# Only colour strips in x-direction
strip <- strip_themed(background_x = elem_list_rect(fill = c("#756BB1", "#21918c","#F0E442"), alpha=c(0.5)))

# Necessary to put RH% into the facet labels
period_names <- as_labeller(
     c(`dry 2022-1` = "Dry 2022-1", `dry 2022-2` = "Dry 2022-2",`wet 2023` = "Wet 2023"))
### plot path similarity inbound and outbound commutes
dtw_seasons <- ggplot(aes(x=direc,y = Distance, color=sameid, group = sameid), data=distsim_all) +
  #geom_boxplot(position=position_dodge(1))+
  geom_jitter(aes(y = Distance, alpha=sameid), position = position_jitterdodge(jitter.width = 0.3, jitter.height = 0.05))+
  #scale_color_manual(values = c("#440154", "#21918c", "#fde725"))+
  scale_alpha_manual(values = c(0.8,0.1))+
  scale_color_manual(values = c("black","grey"))+
  guides(color="none")+
  #scale_alpha_manual(values=c(0.1,0.8))+
  facet_wrap2(.~year_cave, strip=strip, labeller = period_names)+
  stat_summary(aes(y = Distance, group = sameid), fun = mean, 
               color="red",
               position = position_dodge(.8)) +#FF9999
  stat_summary(aes(y = Distance, group = sameid), 
               fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    #color="black",
    color="red",
    width = .3, 
    position = position_dodge(.8)
  ) +
  ylab("DTW pairwise distances")+
  xlab("Path direction")+
  theme_classic()+ 
  theme(legend.position = "none", axis.title = element_text(size=20), axis.title.x = element_text(size = 20), axis.title.y=element_text(size=20), axis.text.y = element_text(size=14), axis.text.x = element_text(size=14), strip.text.x = element_text(size = 12, color="black"))

dtw_seasons

ggsave(file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/inoutflight_simi_plot3_colors.pdf", width=10, height = 5)


dtw_seasons / (outin_id + tracks)+ plot_annotation(tag_levels = "A") &
  theme(
    plot.tag = element_text(size = 20)  # Customize tag size and appearance
  )
  

ggsave(file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/DTW_joined.pdf", width=12, height = 12)
```

### Plot by individual the outbound commutes

```{r}
# plot outbound similarity between and within ids
distsim_all %>% 
  filter(sameid=="yes") %>% 
  #filter(id1!="53506935", id1!= "22507B4E", id1!="38500437", id1!="38506D37") %>%
  ggplot(aes(x=id1, y=Distance, color=direc)) +
  #geom_jitter(aes(alpha=0.5))+
stat_summary(aes(y = Distance, group = direc), fun = mean,
    position = position_dodge(.8)) +
  stat_summary(aes(y = Distance, group = direc),
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "black",
    width = .3, 
    position = position_dodge(.8)
  )+ theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
axis.title.x = element_blank())#annotate("text",x = 30, y = 12, label = "R = 0.22*", size = 4

distsim_all %>% 
  filter(sameid=="no") %>% 
  #filter(id1!="53506935", id1!= "22507B4E", id1!="38500437", id1!="38506D37") %>%
  ggplot(aes(x=year_cave, y=Distance, color=direc)) +
  #geom_jitter(aes(alpha=0.5))+
stat_summary(aes(y = Distance, group = direc), fun = mean,
    position = position_dodge(.8)) +
  stat_summary(aes(y = Distance, group = direc),
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "black",
    width = .3, 
    position = position_dodge(.8)
  )+ theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
axis.title.x = element_blank())
```

### Test repeatability in outbound commutes within indviduals with normal lmr
```{r}
outbound <- distsim_all %>%
  filter(direc=="outbound", sameid=="yes", year_cave=="dry 2022-1")%>%
  mutate(date=paste(date1, date2, sep="_"))

outbound$day_year1 <- yday(outbound$date1)
outbound$day_year2 <- yday(outbound$date2)
outbound$time_diff <- abs(outbound$day_year1-outbound$day_year2)


outbound$cave[which(is.na(outbound$cave))] <- "ajcave_2023"
# Check the levels and counts
table(outbound$cave)

# Check for NAs in Distance, date, and id1
sum(is.na(outbound$Distance))
sum(is.na(outbound$date))
sum(is.na(outbound$id1))

# Convert id1 to factor if necessary
outbound$id1 <- as.factor(outbound$id1)

library(lme4)
# Remove rows where the character in the 'character' column appears only once
outbound_df <- outbound %>%
  group_by(date) %>%
  filter(n() > 1) %>%
  ungroup()


### only for outbound commute
# out_season <- lmer(log(Distance) ~ cave + day_year1+day_year2 + (1|id1) +(1|day_year1/cave)++(1|day_year2/cave), data = outbound)


out_season <- lmer(log(Distance) ~ time_diff +(1|id1) , data = outbound)

summary(out_season)
car::Anova(out_season)
plot(out_season)
qqnorm(residuals(out_season))

library(DHARMa)
simulationOutput <- simulateResiduals(fittedModel = out_season, plot = F)
plot(simulationOutput)

library(MuMIn)
r.squaredGLMM(out_season)
#Rm= This means that % of the variance is explained by the fixed effects
#Rc=This means that 7% of the variance is explained by both fixed and random effects

#calculate repeatability 
#The repeatability value ranges between 0 and 1 and indicates the proportion of variance explained by a given
#random eﬀect, after controlling the model for the fixed eﬀects of time and cave
print(VarCorr(out_season),comp = c("Variance","Std.Dev."))
# individual repeatability
0.087966/ 0.214213
0.10199/0.18372

set.seed(1)
library(arm)
simulated <- sim(out_season, n.sim = 1000)
posterior_animal_id <- apply(simulated@ranef$"id1"[ , , 1],1,var)

#posterior_cave <- apply(simulated@ranef$"year_cave"[ , , 1],1,var)

posterior_residual <- simulated@sigma^2

quantile(posterior_animal_id/(posterior_residual),prob=c(0.025, 0.5, 0.975))
#Alternatively to repeatability we can also calculate the coeﬃcient of variation for between individual variance(CVi)
CVi <- sqrt(posterior_animal_id) / summary(out_season)$coefficients[1]
quantile(CVi,prob=c(0.025, 0.5, 0.975))

library(merTools)
randomSims <- REsim(out_season, n.sims = 1000)
head(randomSims[randomSims$groupFctr=="id1",])
#add cave to the individuals

randomSims <- merge(randomSims[randomSims$groupFctr=="id1",],
outbound[!duplicated(outbound$id1),c("id1","year_cave")],
by.x = "groupID",by.y="id1")

#add mean by cave
randomSims[randomSims$year_cave== "dry 2022-1",]$mean <- randomSims[randomSims$year_cave== "dry 2022-1",]$mean +
fixef(out_season)["(Intercept)"]
randomSims[randomSims$year_cave== "dry 2022-2",]$mean <- randomSims[randomSims$year_cave== "dry 2022-2",]$mean +
(fixef(out_season)["(Intercept)"] + fixef(out_season)["year_cavedry 2022-2"])
randomSims[randomSims$year_cave== "wet 2023",]$mean <- randomSims[randomSims$year_cave== "wet 2023",]$mean +
(fixef(out_season)["(Intercept)"] + fixef(out_season)["year_cavewet 2023"])

# sort data
randomSims$groupID <- factor(randomSims$groupID,
levels = randomSims$groupID[order(randomSims$mean)])

ggplot()+
geom_errorbar(data = randomSims,
aes(x = groupID, ymin = exp(mean-sd)/1000,
ymax = exp(mean+sd)/1000, ))+
geom_point(data = randomSims,
aes(x = groupID, y = exp(mean)/1000), shape = 21, size = 2)+#, fill = year_cave
theme_classic()+
ylab("DTW similarity (km)")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
axis.title.x = element_blank())

ggsave(file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/indvar_dtw_out.pdf", width=12, height = 7)
```

### plot individuals with high and low repeatability
```{r}
p_high_variation <- febpaths %>%
  filter(tag_local_identifier=="2C500235_D" | tag_local_identifier=="37507537_D" | tag_local_identifier=="38506737_G",  direc=="outbound") %>%#"38506737_G"
  ggplot(aes(x=location_long, y=location_lat, color=tag_local_identifier, linetype =direc, sgroup=ID))+
  #scale_color_manual(values =c("#0072B2","#D55E00"))+
  geom_path()+
  xlab("Longitude")+
  ylab("Latitude")+
  theme_classic()+
  theme(axis.title = element_text(face = 'bold', size=20), axis.title.x = element_text(size = 20), axis.title.y=element_text(size=20), axis.text.y = element_text(size=14), axis.text.x = element_text(size=14))

p_low_variation <- febpaths %>%
  filter(tag_local_identifier=="23506B4E_G"| tag_local_identifier=="2C507735_G",  direc=="outbound") %>%
  ggplot(aes(x=location_long, y=location_lat, color=tag_local_identifier, linetype =direc, sgroup=ID))+
  #scale_color_manual(values =c("#0072B2","#D55E00"))+
  geom_path()+
  xlab("Longitude")+
  ylab("Latitude")+
  theme_classic()+
  theme(axis.title = element_text(face = 'bold', size=20), axis.title.x = element_text(size = 20), axis.title.y=element_text(size=20), axis.text.y = element_text(size=14), axis.text.x = element_text(size=14))

p_high_variation + p_low_variation 
```

### Calculate repeatability for outbound commutes with Rptr
```{r}
library(rptR)
###Repeatability of the model
#split by cave and calculate repeatability by population
out_rep<- rpt(log(Distance)~time_diff + (1|id1), grname = c("id1","Residual","Fixed"), 
    data = outbound, datatype = "Gaussian", nboot = 1000, npermut = 0, ratio = FALSE, adjusted = FALSE)

print(out_rep)
0.102/(0.184 +0.02)
0.045/(0.011 +0.002)

plot(out_rep, grname = "id1", scale = "probit")

####
# Interindividual variance is the variance of the random effects (random effects variance of 1|id)
# Residual variance of the random effects (within-individual variance)
### Example
# Groups   Name        Std.Dev.
 #id       (Intercept)  1.2 # interindividual variance
 #Residual              0.8 # within individual variance

# Extract random effects (BLUPs) for 'id1' from the model
random_effects <- ranef(out_rep$mod)$id1

# Calculate the variance of the random effects to estimate interindividual variance
interindividual_variance <- var(random_effects[, 1], na.rm = TRUE)
print(interindividual_variance)

# Extract residuals from the model
model_residuals <- residuals(out_rep$mod)

# Calculate the total residual variance (within-individual variance)
residual_variance <- var(model_residuals, na.rm = TRUE)
print(residual_variance)

# Calculate population-level repeatability (R)
R_population <- interindividual_variance / (interindividual_variance + residual_variance)
print(R_population)

library(dplyr)
#index of individual repeatabilitu
# Add residuals to the original data
# Extract residuals from the model
model_residuals <- residuals(out_rep$mod)
outbound$residuals <- model_residuals
mean_res <- mean(outbound$residuals, na.rm=TRUE)

# Calculate intraindividual variance for each individual (within-individual variance per individual)
intra_variance <- outbound %>%
  group_by(id1) %>%
  dplyr::summarise(S_ind =mean(residuals - mean_res,na.rm = TRUE)^2)

print(intra_variance)

# Calculate individual-level repeatability for each individual
intraindividual_repeatability <- intra_variance %>%
  mutate(R_ind = interindividual_variance / (interindividual_variance + S_ind))

# Print the individual-level repeatability values
print(intraindividual_repeatability)
mean(intraindividual_repeatability$R_ind, na.rm=TRUE)
sd(intraindividual_repeatability$R_ind, na.rm=TRUE)
```

Interindividual variance  is estimated from the variance of the random effects for id1.
Within-individual variance is estimated from the variance of the residuals of the model.
Intraindividual variance is calculated as the variance of residuals for each individual.
Population-level repeatability and individual-level repeatability are calculated using these estimates


### Test repeatability in inbound commutes within and between individuals with lme

```{r}
inbound <- distsim_all %>%
  filter(direc=="inbound", sameid=="yes", year_cave=="dry 2022-1")%>%
  mutate(date=paste(date1, date2, sep="_"))

inbound$day_year1 <- yday(inbound$date1)
inbound$day_year2 <- yday(inbound$date2)
inbound$time_diff <- abs(inbound$day_year1-inbound$day_year2)


# Check the levels and counts
table(inbound$year_cave)

# Check for NAs in Distance, date, and id1
sum(is.na(inbound$Distance))
sum(is.na(inbound$date))
sum(is.na(inbound$id1))

# Convert id1 to factor if necessary
inbound$id1 <- as.factor(inbound$id1)

### only for outbound commute
# out_season <- lmer(log(Distance) ~ cave + day_year1+day_year2 + (1|id1) +(1|day_year1/cave)++(1|day_year2/cave), data = outbound)
in_season <- lmer(log(Distance) ~  time_diff +(1|id1), data = inbound)

summary(in_season)
plot(in_season)
qqnorm(residuals(in_season))
car::Anova(in_season)

acf(residuals(in_season), main = "ACF of Residuals (With no Autocorrelation)")

simulationOutput <- simulateResiduals(fittedModel = in_season, plot = F)
plot(simulationOutput)
library(MuMIn)
r.squaredGLMM(in_season)
#Rm= This means that % of the variance is explained by the fixed effects
#Rc=This means that 7% of the variance is explained by both fixed and random effects

#calculate repeatability 
#The repeatability value ranges between 0 and 1 and indicates the proportion of variance explained by a given
#random eﬀect, after controlling the model for the fixed eﬀects of month and sex in our case
print(VarCorr(in_season),comp = c("Variance","Std.Dev."))
# individual repeatability
0.094215/0.115265

set.seed(1)
library(arm)

simulated <- sim(in_season, n.sim = 1000)
posterior_animal_id <- apply(simulated@ranef$"id1"[ , , 1],1,var)
#posterior_animal_id_year <- apply(simulated@ranef$"year_cave:id1"[ , , 1],1,var)
posterior_residual <- simulated@sigma^2

quantile(posterior_animal_id/c(posterior_residual),prob=c(0.025, 0.5, 0.975), na.rm=TRUE)
#Alternatively to repeatability we can also calculate the coeﬃcient of variation for between individual variance(CVi)
CVi <- sqrt(posterior_animal_id) / summary(in_season)$coefficients[1]
quantile(CVi,prob=c(0.025, 0.5, 0.975))

library(merTools)
randomSims <- REsim(in_season, n.sims = 1000)
head(randomSims[randomSims$groupFctr=="id1",])
#add cave to the individuals
randomSims <- merge(randomSims[randomSims$groupFctr=="id1",],
inbound[!duplicated(inbound$id1),c("id1","year_cave")],
by.x = "groupID",by.y="id1")

#add mean by cave
randomSims[randomSims$year_cave== "dry 2022-1",]$mean <- randomSims[randomSims$year_cave== "dry 2022-1",]$mean +
fixef(in_season)["(Intercept)"]
randomSims[randomSims$year_cave== "dry 2022-2",]$mean <- randomSims[randomSims$year_cave== "dry 2022-2",]$mean +
(fixef(in_season)["(Intercept)"] + fixef(in_season)["year_cavedry 2022-2"])
randomSims[randomSims$year_cave== "wet 2023",]$mean <- randomSims[randomSims$year_cave== "wet 2023",]$mean +
(fixef(in_season)["(Intercept)"] + fixef(in_season)["year_cavewet 2023"])

# sort data
randomSims$groupID <- factor(randomSims$groupID,
levels = randomSims$groupID[order(randomSims$mean)])

ggplot()+
geom_errorbar(data = randomSims,
aes(x = groupID, ymin = exp(mean-sd)/1000,
ymax = exp(mean+sd)/1000, ))+
geom_point(data = randomSims,
aes(x = groupID, y = exp(mean)/1000), shape = 21, size = 2)+#fill = year_cave
theme_classic()+
ylab("DTW similarity (km)")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
axis.title.x = element_blank())

ggsave(file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/indvar_dtw_in.pdf", width=12, height = 7)

in_high_variation <- febpaths %>%
  filter(tag_local_identifier=="37506237_D" | tag_local_identifier=="38506737_G",  direc=="inbound") %>%
  ggplot(aes(x=location_long, y=location_lat, color=tag_local_identifier, linetype =direc, sgroup=ID))+
  #scale_color_manual(values =c("#0072B2","#D55E00"))+
  geom_path()+
  xlab("Longitude")+
  ylab("Latitude")+
  theme_classic()+
  theme(axis.title = element_text(face = 'bold', size=20), axis.title.x = element_text(size = 20), axis.title.y=element_text(size=20), axis.text.y = element_text(size=14), axis.text.x = element_text(size=14))

in_low_variation <- febpaths %>%
  filter(tag_local_identifier=="23506B4E_G"| tag_local_identifier=="2E500235_G",  direc=="inbound") %>%
  ggplot(aes(x=location_long, y=location_lat, color=tag_local_identifier, linetype =direc, sgroup=ID))+
  #scale_color_manual(values =c("#0072B2","#D55E00"))+
  geom_path()+
  xlab("Longitude")+
  ylab("Latitude")+
  theme_classic()+
  theme(axis.title = element_text(face = 'bold', size=20), axis.title.x = element_text(size = 20), axis.title.y=element_text(size=20), axis.text.y = element_text(size=14), axis.text.x = element_text(size=14))

in_high_variation + in_low_variation 
```

### Repeatability of the inbound commutes with Rpt

```{r}
in_rep<- rpt(log(Distance)~ time_diff + (1|id1), grname = c("id1","Residual","Fixed"), 
    data = inbound, datatype = "Gaussian", nboot = 1000, npermut = 0, ratio = FALSE, adjusted = FALSE)

print(in_rep)

plot(in_rep, grname = "id1", scale = "probit")

#within-individual variance
0.094/(0.115+0.001)
0.039/(0.006+0.001)

#index of individual repeatabilitu
# Add residuals to the original data
# Extract residuals from the model
model_residuals <- residuals(in_rep$mod)
inbound$residuals <- model_residuals
mean_res <- mean(inbound$residuals, na.rm=TRUE)

# Calculate intraindividual variance for each individual (within-individual variance per individual)
intraindividual_variance <- inbound %>%
  group_by(id1) %>%
  dplyr::summarise(S_ind =mean(residuals - mean_res,na.rm = TRUE)^2)

print(intraindividual_variance)

# Calculate individual-level repeatability for each individual
intraindividual_repeatability <- intraindividual_variance %>%
  mutate(R_ind = interindividual_variance / (interindividual_variance + S_ind))

# Print the individual-level repeatability values
print(intraindividual_repeatability)
mean(intraindividual_repeatability$R_ind, na.rm=TRUE)
sd(intraindividual_repeatability$R_ind, na.rm=TRUE)
```

### Test how inbound and outbound commute changes between periods based on sameid

```{r}
# out_all <- distsim_all %>%
#   filter(direc=="outbound")

hist(log(distsim_all$Distance))
library(lmerTest)
m_path <- lmer(log(Distance) ~ year_cave*sameid*direc + (1|dyad), data = distsim_all)

summary(m_path)
plot(m_path)
car::Anova(m_path)

# Extract the model summary
model_summary <- summary(m_path)

# Extract t-values from the fixed effects
t_values <- model_summary$coefficients[, "t value"]

# Calculate p-values using the normal distribution
p_values <- 2 * (1 - pnorm(abs(t_values)))

# Create a table with the effect names, t-values, and p-values
effects_table <- data.frame(
  Effect = rownames(model_summary$coefficients),
  `t value` = t_values,
  `p value` = p_values
)

# Print the table
print(effects_table)

library(MuMIn)
r.squaredGLMM(m_path)

meanwithinbwteen <- distsim_all%>%
  group_by(sameid)%>%
  reframe(meandis=mean(Distance), sddist=sd(Distance))

meandirec <- distsim_all%>%
  group_by(direc)%>%
  reframe(meandis=mean(Distance), sddist=sd(Distance))


#random slope for id
# m_out_bp_rs <- lmer(log(Distance) ~ year_cave*sameid + (1|dyad), data = out_all)
# summary(m_out_bp_rs)
# 
# car::Anova(m_out_bp_rs)
# 
# simulationOutput <- simulateResiduals(fittedModel = m_out_bp_rs, plot = F)
# plot(simulationOutput)
# 
# AIC(m_out_bp, m_out_bp_rs)

#filter inbound commute
# in_all <- distsim_all %>%
#   filter(direc=="inbound")
# 
# m_in_bp <- lmer(log(Distance) ~ year_cave*sameid + (1|dyad), data = in_all)
# summary(m_in_bp)
# 
# car::Anova(m_in_bp)
# 
# simulationOutput <- simulateResiduals(fittedModel = m_in_bp, plot = F)
# plot(simulationOutput)
# 
# 
# m_in_bp_rs <- lmer(log(Distance) ~ year_cave+sameid + (1|dyad), data = in_all)
# summary(m_in_bp_rs)
# 
# car::Anova(m_in_bp_rs)
# 
# simulationOutput <- simulateResiduals(fittedModel = m_in_bp_rs, plot = F)
# plot(simulationOutput)
# 
# AIC(m_in_bp, m_in_bp_rs)


```



### Difference in pairwise wind speed and pairwise DTW within individuals

```{r}
# read wind data
windall_df <- read.csv(file="/Users/ccalderon/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/data/allpaths_winddata.csv")
distsim_all <- read.csv(file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/data/distsimtable_commutes.csv")


windall_df$wind_dire_rad <- windall_df$wind_dire*(pi/180)
# calculate mean by ids
meanwind_id <- windall_df %>%
  mutate(date=date(timestamp))%>%
  group_by(ID, tag.local.identifier, date, direc)%>%
  dplyr::summarise(meanwind=mean(wind.speed),varwind=var(wind.speed), meanwd=mean(wind_dire_rad),varwd=var(wind_dire_rad))


meanwind_id_ls <- split(meanwind_id, meanwind_id$direc)

# Create all pairwise combinations of dataframes for INBOUND commutes
pairs_windin<- expand.grid(meanwind_id_ls[[1]]$ID, meanwind_id_ls[[1]]$ID)
pairs_windin <- subset(pairs_windin, Var1 != Var2)

# Rename columns in pairs_windout for clarity
names(pairs_windin) <- c("Pair1", "Pair2")

# Merge pairs_windout with the original data twice, for ID1 and ID2
pairs_windin <- merge(pairs_windin, meanwind_id_ls[[1]], by.x = "Pair1", by.y = "ID", suffixes = c("_Pair1", ""))
pairs_windin <- merge(pairs_windin, meanwind_id_ls[[1]], by.x = "Pair2", by.y = "ID", suffixes = c("_Pair1", "_Pair2"))

# Calculate the absolute difference in meanwind for each pair
pairs_windin$windspDf <- abs(pairs_windin$meanwind_Pair1 - pairs_windin$meanwind_Pair2)
pairs_windin$windspVar <- abs(pairs_windin$varwind_Pair1 - pairs_windin$varwind_Pair2)
pairs_windin$meanwd <- abs(pairs_windin$meanwd_Pair1 - pairs_windin$meanwd_Pair2)
pairs_windin$varwd <- abs(pairs_windin$varwd_Pair1 - pairs_windin$varwd_Pair2)

colnames(pairs_windin)[5] <- "direc"


# Create all pairwise combinations of dataframes for OUTBOUND commutes
pairs_windout<- expand.grid(meanwind_id_ls[[2]]$ID, meanwind_id_ls[[2]]$ID)
pairs_windout <- subset(pairs_windout, Var1 != Var2)

# Rename columns in pairs_windout for clarity
names(pairs_windout) <- c("Pair1", "Pair2")

# Merge pairs_windout with the original data twice, for ID1 and ID2
pairs_windout <- merge(pairs_windout, meanwind_id_ls[[2]], by.x = "Pair1", by.y = "ID", suffixes = c("_Pair1", ""))
pairs_windout <- merge(pairs_windout, meanwind_id_ls[[2]], by.x = "Pair2", by.y = "ID", suffixes = c("_Pair1", "_Pair2"))

# Calculate the absolute difference in meanwind for each pair
pairs_windout$windspDf <- abs(pairs_windout$meanwind_Pair1 - pairs_windout$meanwind_Pair2)
pairs_windout$windspVar <- abs(pairs_windout$varwind_Pair1 - pairs_windout$varwind_Pair2)
pairs_windout$meanwd <- abs(pairs_windout$meanwd_Pair1 - pairs_windout$meanwd_Pair2)
pairs_windout$varwd <- abs(pairs_windout$varwd_Pair1 - pairs_windout$varwd_Pair2)
colnames(pairs_windout)[5] <- "direc"

# Merge the two data frames
pairs_wind <- rbind(pairs_windout,pairs_windin)

# Merge distsim_all with pairs_wind on Pair1 and Pair2
distsim_all <- merge(distsim_all, pairs_wind[, c("Pair1", "Pair2","direc", "windspDf", "windspVar", "meanwd","varwd")], 
                     by = c("Pair1", "Pair2", "direc"), all.x = TRUE)


### calculate crosswind and windsupport
meanws <- windall_df %>%
    mutate(date=date(timestamp))%>%
    filter(date<"2022-02-28")%>%
    group_by(direc)%>%
    dplyr::summarise(meancw=mean(cw, na.rm=TRUE), meanws=mean(ws, na.rm=TRUE), sdws=sd(ws, na.rm=TRUE), varcw=var(cw, na.rm=TRUE), varws=var(ws, na.rm=TRUE))

# calculate mean crosswind by ids
meancw_ws <- windall_df %>%
  mutate(date=date(timestamp))%>%
  group_by(ID, tag.local.identifier, date, direc)%>%
  dplyr::summarise(meancw=mean(cw), meanws=mean(ws), varcw=var(cw), varws=var(ws))%>%
  filter(date<"2022-02-28")


meancw_ws_id_ls <- split(meancw_ws, meancw_ws$direc)

# Create all pairwise combinations of dataframes for INBOUND commutes
pairs_cwws_in<- expand.grid(meancw_ws_id_ls[[1]]$ID, meancw_ws_id_ls[[1]]$ID)
pairs_cwws_in <- subset(pairs_cwws_in, Var1 != Var2)

# Rename columns in pairs_windout for clarity
names(pairs_cwws_in) <- c("Pair1", "Pair2")

# Merge pairs_windout with the original data twice, for ID1 and ID2
pairs_cwws_in <- merge(pairs_cwws_in, meancw_ws_id_ls[[1]], by.x = "Pair1", by.y = "ID", suffixes = c("_Pair1", ""))
pairs_cwws_in <- merge(pairs_cwws_in, meancw_ws_id_ls[[1]], by.x = "Pair2", by.y = "ID", suffixes = c("_Pair1", "_Pair2"))

# Calculate the absolute difference in meanwind for each pair
pairs_cwws_in$difCW <- abs(pairs_cwws_in$meancw_Pair1 - pairs_cwws_in$meancw_Pair2)
pairs_cwws_in$difWS <- abs(pairs_cwws_in$meanws_Pair1 - pairs_cwws_in$meanws_Pair2)
pairs_cwws_in$varCW <- abs(pairs_cwws_in$varcw_Pair1 - pairs_cwws_in$varcw_Pair2)
pairs_cwws_in$varWS <- abs(pairs_cwws_in$varws_Pair1 - pairs_cwws_in$varws_Pair2)
colnames(pairs_cwws_in)[5] <- "direc"


# Create all pairwise combinations of dataframes for OUTBOUND commutes
pairs_cwws_out<- expand.grid(meancw_ws_id_ls[[2]]$ID, meancw_ws_id_ls[[2]]$ID)
pairs_cwws_out <- subset(pairs_cwws_out, Var1 != Var2)

# Rename columns in pairs_windout for clarity
names(pairs_cwws_out) <- c("Pair1", "Pair2")

# Merge pairs_windout with the original data twice, for ID1 and ID2
pairs_cwws_out <- merge(pairs_cwws_out, meancw_ws_id_ls[[2]], by.x = "Pair1", by.y = "ID", suffixes = c("_Pair1", ""))
pairs_cwws_out <- merge(pairs_cwws_out, meancw_ws_id_ls[[2]], by.x = "Pair2", by.y = "ID", suffixes = c("_Pair1", "_Pair2"))

# Calculate the absolute difference in meanwind for each pair
pairs_cwws_out$difCW <- abs(pairs_cwws_out$meancw_Pair1 - pairs_cwws_out$meancw_Pair2)
pairs_cwws_out$difWS <- abs(pairs_cwws_out$meanws_Pair1 - pairs_cwws_out$meanws_Pair2)
pairs_cwws_out$varCW <- abs(pairs_cwws_out$varcw_Pair1 - pairs_cwws_out$varcw_Pair2)
pairs_cwws_out$varWS <- abs(pairs_cwws_out$varws_Pair1 - pairs_cwws_out$varws_Pair2)
colnames(pairs_cwws_out)[5] <- "direc"

# Merge the two data frames
pairs_cwws_inout <- rbind(pairs_cwws_out,pairs_cwws_in)

# Merge distsim_all with pairs_wind on Pair1 and Pair2
distsim_all <- merge(distsim_all, pairs_cwws_inout[, c("Pair1", "Pair2","direc", "difCW","difWS", "varCW","varWS")], 
                     by = c("Pair1", "Pair2", "direc"), all.x = TRUE)


distsim_all_sameid <- distsim_all %>%
  filter(sameid=="yes")
 
# Plot mean wind speed
winddif_dist <- ggplot(aes(x=windspDf, y=log(Distance)),data=distsim_all_sameid)+
  geom_point()+
  theme_classic()+
  ylab("Pairwise distance within \n individual commutes")+
  xlab("Mean wind speed \n difference (m/s)")+
 theme(legend.position = "none", 
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18), 
        plot.title= element_text(size = 20))+
  facet_wrap(.~direc)+
  geom_smooth(method="lm")

winddif_var <- ggplot(aes(x=windspVar, y=log(Distance)),data=distsim_all_sameid)+
  geom_point()+
  theme_classic()+
  ylab("Pairwise DTW distance of \n commutes within ids")+
  xlab("Wind speed variance \n difference (m/s)")+
  theme(legend.position = "none", 
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18), 
        plot.title= element_text(size = 20))+
  facet_wrap(.~direc)+
  geom_smooth(method="lm")

### wind direction
# winddirec_mean <- ggplot(aes(x=meanwd, y=log(Distance)),data=distsim_all_sameid)+
#   geom_point()+
#   theme_classic()+
#   ylab("Pairwise DTW distance of \n commutes within ids")+
#   xlab("Mean wind direction \n difference (rad)")+
#   theme(axis.title = element_text(face = 'bold', size=20), 
#         axis.title.x = element_text(size = 24), 
#         axis.title.y=element_text(size=24), 
#         axis.text.y = element_text(size=20), 
#         axis.text.x = element_text(size=20),
#         legend.title =element_text(size=20),
#         legend.text = element_text(size=18)
#         )+
#   facet_wrap(.~direc)+
#   geom_smooth(method="lm", color="black")
# 
# winddirec_var <- ggplot(aes(x=varwd, y=log(Distance)),data=distsim_all_sameid)+
#   geom_point()+
#   theme_classic()+
#   ylab("Pairwise DTW distance of \n commutes within ids")+
#   xlab("Variance wind direction \n difference (rad)")+
#   theme(axis.title = element_text(face = 'bold', size=20), 
#         axis.title.x = element_text(size = 24), 
#         axis.title.y=element_text(size=24), 
#         axis.text.y = element_text(size=20), 
#         axis.text.x = element_text(size=20),
#         legend.title =element_text(size=20),
#         legend.text = element_text(size=18)
#         )+
#   facet_wrap(.~direc)+
#   geom_smooth(method="lm", color="black")

# Plot cross wind
cwdiff_dist <-ggplot(aes(x=difCW, y=log(Distance)),data=distsim_all_sameid)+
  geom_point()+
  theme_classic()+
  ylab("Pairwise DTW distance of \n commutes within ids")+
  xlab("Mean crosswind \n difference (m/s)")+
 theme(legend.position = "none", 
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18), 
        plot.title= element_text(size = 20))+
  facet_wrap(.~direc)+
  geom_smooth(method="lm")


cwvar_dist <-ggplot(aes(x=varCW, y=log(Distance)),data=distsim_all_sameid)+
  geom_point()+
  theme_classic()+
  ylab("Pairwise DTW distance of \n commutes within ids")+
  xlab("Crosswind variance \n difference (m/s)")+
 theme(legend.position = "none", 
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18), 
        plot.title= element_text(size = 20))+
  facet_wrap(.~direc)+
  geom_smooth(method="lm")


# Plot wind support
wsdiff_dist <- ggplot(aes(x=difWS, y=log(Distance)),data=distsim_all_sameid)+
  geom_point()+
  theme_classic()+
  ylab("Pairwise DTW distance of \n commutes within ids")+
  xlab("Mean wind support \n difference (m/s)")+
  theme(legend.position = "none", 
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18), 
        plot.title= element_text(size = 20))+
  facet_wrap(.~direc)+
  geom_smooth(method="lm")
  

wsvar_dist <-ggplot(aes(x=varWS, y=log(Distance)),data=distsim_all_sameid)+
  geom_point()+
  theme_classic()+
  ylab("Pairwise DTW distance of \n commutes within ids")+
  xlab("Wind support variance \n difference (m/s)")+
  theme(legend.position = "none", 
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18), 
        plot.title= element_text(size = 20))+
  facet_wrap(.~direc)+
  geom_smooth(method="lm")

library(patchwork)
winddif_dist 

# mean plot wind variables
wsdiff_dist / cwvar_dist + plot_layout(axes="collect_y") + plot_annotation(tag_levels = "A")

ggsave(file="/Users/ccalderon/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/figures/MeanVar_WS_CW_DTW.pdf", width=7, height=7)

#variance plot mean variable
wsvar_dist/cwvar_dist  + plot_layout(axes="collect_y")

ggsave(file="/Users/ccalderon/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/figures/Var_WS_CW_DTW.pdf", width=7, height=12)

# variance in crosswind can have an affect on both if they change more shape on the inbound commutes when they cross the ocean


(wsdiff_dist + cwvar_dist)/ combined_plot 

```

### Test that pairwise distance have no relationship with wind
```{r}
#Test with a linear regression

cor_matrix <- cor(distsim_all_sameid[, c("windspDf", "difCW", "difWS", "windspVar", "varCW", "varWS")], use = "complete.obs")
print(cor_matrix)


names(distsim_all_sameid)
### Check this better
model_path_wsp_mean <- lmer(log(Distance)~windspDf +difCW +direc + (1|dyad), data=distsim_all_sameid)

summary(model_path_wsp_mean)
car::Anova(model_path_wsp_mean)
plot(model_path_wsp_mean)

model_path_ws <- lmer(log(Distance)~difCW + difWS+direc + (1|dyad), data=distsim_all_sameid)

summary(model_path_ws)
car::Anova(model_path_ws)
plot(model_path_ws)

# Calculate marginal means for 'direc'
em_direc <- emmeans(model_path_ws, ~ direc)
pairwise_em_direc <- pairs(em_direc)

model_path_ws_var <- lmer(log(Distance)~ windspVar+varCW+varWS+direc+ (1|dyad), data=distsim_all_sameid)

summary(model_path_ws_var)
car::Anova(model_path_ws_var)
plot(model_path_ws_var)  


AIC(model_path_wsp_mean,model_path_ws,model_path_ws_var)

```
### Figure of distnac evs wind with predicted values of the model
```{r}
# Generate prediction data
new_data <- expand.grid(
  difWS = seq(min(distsim_all_sameid$difWS, na.rm=TRUE), max(distsim_all_sameid$difWS, na.rm=TRUE), length.out = 100),
  difCW = mean(distsim_all_sameid$difCW, na.rm=TRUE),  # Hold difWS constant
  direc = unique(distsim_all_sameid$direc)  # Include both "outbound" and "inbound"
)

new_data$predicted_log_distance <- predict(model_path_ws, newdata = new_data, re.form = NA)
new_data$predicted_se <- predict(model_path_ws, newdata = new_data, re.form = NA, se.fit = TRUE)$se.fit
new_data$lower <- new_data$predicted_log_distance - 1.96 * new_data$predicted_se
new_data$upper <- new_data$predicted_log_distance + 1.96 * new_data$predicted_se


# Plot predictions
# Add observed values of log(Distance)
distsim_all_sameid$log_distance <- log(distsim_all_sameid$Distance)
#distsim_all_sameid$direc_f <- factor(distsim_all_sameid$direc, levels = c("outbound","inbound"))

# Plot predictions with confidence intervals and observed data
ws_dist <- ggplot() +
  geom_ribbon(data = new_data, aes(x = difWS, ymin = lower, ymax = upper), alpha = 0.2) +
  geom_line(data = new_data, aes(x = difWS, y = predicted_log_distance), color="#619CFF", size = 1) +
  geom_point(data = distsim_all_sameid, aes(x = difWS, y = log_distance), alpha = 0.6) +
  scale_y_continuous(breaks = c(9, 10, 11), labels = c("8100", "22000", "59800"))+
  labs(
    x = "Mean wind support (m/s)",
    y = "Predicted log pairwise DTW distance of \n commutes within ids "
  ) +
  guides(fill="none", color="none")+
  theme(legend.position = "none", 
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18), 
        plot.title= element_text(size = 20))+
  theme_classic()+
  facet_wrap(.~direc,scales = "free")

#### Calculate the predicted values for cross wind
# Generate prediction data
new_data_2 <- expand.grid(
  varCW = seq(min(distsim_all_sameid$varCW, na.rm=TRUE), max(distsim_all_sameid$varCW, na.rm=TRUE), length.out = 100),
  varWS = mean(distsim_all_sameid$varWS, na.rm=TRUE),
  windspVar = mean(distsim_all_sameid$windspVar, na.rm=TRUE),# Hold difWS constant
  direc = unique(distsim_all_sameid$direc)  # Include both "outbound" and "inbound"
)

new_data_2$predicted_log_distance <- predict(model_path_ws_var, newdata = new_data_2, re.form = NA)
new_data_2$predicted_se <- predict(model_path_ws_var, newdata = new_data_2, re.form = NA, se.fit = TRUE)$se.fit
new_data_2$lower <- new_data_2$predicted_log_distance - 1.96 * new_data_2$predicted_se
new_data_2$upper <- new_data_2$predicted_log_distance + 1.96 * new_data_2$predicted_se

# Add observed values of log(Distance)
distsim_all_sameid$log_distance_2 <- log(distsim_all_sameid$Distance)

# Plot predictions with confidence intervals and observed data
cw_dist <- ggplot() +
  geom_ribbon(data = new_data_2, aes(x = varCW, ymin = lower, ymax = upper), alpha = 0.2) +
  geom_line(data = new_data_2, aes(x = varCW, y = predicted_log_distance), color="#619CFF", size = 1) +
  geom_point(data = distsim_all_sameid, aes(x = varCW, y = log_distance_2), alpha = 0.6) +
  scale_y_continuous(breaks = c(9, 10, 11), labels = c("8100", "22000", "59800"))+
  labs(
    x = "Crosswind variance \n difference (m/s)",
    y = "Predicted log pairwise DTW distance of \n commutes within ids "
  ) +
  guides(fill="none", color="none")+
  theme(legend.position = "none", 
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18), 
        plot.title= element_text(size = 20))+
  theme_classic()+
  facet_wrap(.~direc,scales = "free")

# mean plot wind variables
ws_dist / cw_dist + plot_layout(axes="collect_y") + plot_annotation(tag_levels = "A")

ggsave(file="/Users/ccalderon/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/figures/Fig2.pdf", width=8, height=7)

```

