---
title: "Dynamic time warping"
output:
  pdf_document: default
  html_document:
    df_print: paged
date: "2024-03-27"
Author: Camila
---

In this code, I aim to calculte DTW distance for individuals in the different tracking periods, to see how similar are outbounds and inbounds commutes within and between individuals

### Load libraries

```{r setup, warning=FALSE, message=FALSE}
library(dtw)  # dynamic time warping
library(sp)  # handle spatial objects
library(rgeos)# geoprocessing functions
library(lubridate)
library(dplyr)
library(ggplot2)
library(dtwclust)
library(cluster)    # clustering algorithms
library(factoextra) # clustering algorithms & visualization
library(move)
library(stringr)
library(ggmap)
library(sf)
library(purrr)
```

### Load data

```{r loading, results='hide'}
load(file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/data/bocas_2022-2023HMMbehaviors.RData")

# order data frame
bocas <- bocas[order(bocas$tag_local_identifier, bocas$timestamp),]

bocas_used <- bocas %>%
  filter(year_cave=="2022_lagruta_Feb"|year_cave== "2022_ajcave"|year_cave== "2023_ajcave")

move_path <- move(data=bocas_used,
                  x=bocas_used$location_long, 
                  y=bocas_used$location_lat, 
                  time=as.POSIXct(bocas_used$timestamp, format="%Y-%m-%d %H:%M:%S", tz="UTC"), 
                  animal = bocas_used$ID_batday,
                  sensor = bocas_used$sensor_type,
                  proj = CRS("+proj=longlat +ellps=WGS84"))

# create a movestack
# add movement parameters, pre-cleaning
moveList <- lapply(split(move_path), function(myInd){
  myInd$tlag <- c(NA, move::timeLag(myInd, units="secs"))
  return(myInd)
})

# check time tracked for each id
hourstracked <- lapply(1:length(moveList), function(id){
  #x$tlag[x$tlag>=50000] <- NA
  id <- names(moveList[id])
  s <- sum(moveList[[id]]$tlag, na.rm = TRUE)
  hours <- data.frame(idday=id, h=s/3600)
  return(hours)
})

hourstracked <- do.call(rbind.data.frame,hourstracked)
hourstracked$id <- substr(hourstracked$idday, 1,nchar(hourstracked$idday)-2)

hourssume <- hourstracked %>%
  group_by(id)%>%
  dplyr::summarise(sum_hours= sum(h), meanhours= mean(h), count = n())

# convert move object to dataframe
phas <- moveStack(moveList, forceTz="UTC")
bocas_used <- as.data.frame(phas) 
bocas_used$tlag[which(bocas_used$behav=="foraging" & bocas_used$tlag>1800)] <- 190

# sampling effort
samplingeffort <- bocas_used %>%
  group_by(tag_local_identifier)%>%
  dplyr::summarise(hours=sum(tlag, na.rm = TRUE)/3600)

# split dataframe
bocas_ls <- split(bocas, bocas$ID)

```

## Select commuting tracks to the foraging spots from the colony 1 groups in February 2022

```{r, message=FALSE, warning=FALSE, results=FALSE}
# adding a variable that counts the length of consecutive commuting and foraging points. This will be used to filter only commuting tracks that contain more than 5 points in a row
bocas$n <- NA
r <- rle(bocas$behav)
i=1
for(i in 1:length(r$lengths)){
  #if(r$values[i] == "commuting"){
      start <- {}
      end <- {}
      try(start <- sum(r$lengths[1:(i-1)])+1)
      if(i == 1) start <- 1
      end <- sum(r$lengths[1:i])

      bocas$n[start:end] <- r$lengths[i]
}

# filter only commuting with less than 5 points in a row
bats_commu <- bocas %>% 
  filter(behav=="commuting" & n>=4 & location_long<=-80)

# add time difference to split commuting tracks by the ID_day
bats_commu <- bats_commu %>%
  group_by(ID_batday) %>%
  mutate(diff = timestamps - lag(timestamps))
bats_commu$diff <- as.numeric(bats_commu$diff)
bats_commu$diff[is.na(bats_commu$diff)] <- 0

# data frame with data from February 2022
commu_dry <- bats_commu %>% 
 filter(date>"2022-01-01" & date < "2022-03-01") %>%
  filter(tag_local_identifier!= "0C506E35_G")#remove individual with only 2 nights of tracking
#

# splitting tracks by ID
commu.split <- split(commu_dry, f=commu_dry$ID)

#remove the first two rows for these individuals which make the classification weird
commu.split[["2C507735_G_2022-02-01"]] <- commu.split[["2C507735_G_2022-02-01"]][c(-1,-2),]
commu.split[["2C507735_G_2022-02-03"]] <- commu.split[["2C507735_G_2022-02-03"]][c(-1,-2),]

# for loop to divide each day of the animal in commuting outbound and commuting inbound using different threshold values
mylist <- list()
for (i in 1:length(commu.split)){
  x <- which(commu.split[[i]]$diff>=600 & commu.split[[i]]$diff<=601 |commu.split[[i]]$diff>=900 & commu.split[[i]]$diff<=970 | commu.split[[i]]$diff>=1000 & commu.split[[i]]$diff<=1100 | commu.split[[i]]$diff>=1300 & commu.split[[i]]$diff<=1322 | commu.split[[i]]$diff>=1570 & commu.split[[i]]$diff<=2200 | commu.split[[i]]$diff>=2526 & commu.split[[i]]$diff<2527 | commu.split[[i]]$diff>=2402  & commu.split[[i]]$diff<=2765 | commu.split[[i]]$diff>=3123 & commu.split[[i]]$diff<=3728 | commu.split[[i]]$diff>=2800 & commu.split[[i]]$diff<=2900 | commu.split[[i]]$diff>=4000)#70 minutes in seconds  2200 2402 2523 commu.split[[i]]$diff>=482 & commu.split[[i]]$diff<=490 | 
  mylist[[i]] <- split(commu.split[[i]], cumsum(1:nrow(commu.split[[i]])%in%x))
  indv <- unique(names(commu.split[i]))
  names(mylist[[i]]) <- paste(indv, seq_along(mylist[[i]]), sep = "_")
}

# convert nested list to data frame
find_df <- function(x) {
  if (is.data.frame(x))
    return(list(x))
  if (!is.list(x))
    return(NULL)
  unlist(lapply(x, find_df), FALSE)
}

data <- find_df(mylist)

# # add name column to list
f <- function (data, name){
  data$name <- name
  data
}

data <- Map(f, data, names(data))

# convert list to dataframe
data.commu <- as.data.frame(do.call(rbind, data))

# remove ID
data.commu <- data.commu%>%
  filter(ID!="37506237_D_2022-02-01")

data.commu$name[which(data.commu$name=="2D507235_G_2022-02-07_2")] <- "2D507235_G_2022-02-07_1"

data.commu.ls <- split(data.commu, data.commu$name)

lapply(split(data.commu, data.commu$tag_local_identifier), function(x){
 ggplot(aes(x=location_long, y=location_lat, col=ID, group = date), data=x)+
  geom_point()#+
  #theme(legend.position = "none") 
})

# select only outbound commuting
data.commu.out <- data.commu[grep("_1", data.commu$name),]
unique(data.commu.out$name)  


# to check that I selected only outbound commutes
x <- split(data.commu.out, data.commu.out$name)

# plot outbound commutes
lapply(split(data.commu.out, data.commu.out$tag_local_identifier), function(x){
 ggplot(aes(x=location_long, y=location_lat, col=name), data=x)+
  geom_point()#+
  #theme(legend.position = "none") 
})

# select only inbound commuting
data.commu.in <- data.commu[grep("_1", invert=TRUE,data.commu$name),]
unique(data.commu.in$name)  

data.commu.in <- data.commu.in %>%
  group_by(name)%>%
  filter(n>=11 & name!="23506B4E_G_2022-02-01_3")

# to check that I selected only inbound commutes
y <- split(data.commu.in, data.commu.in$name)

# plot inbound commutes
lapply(split(data.commu.in, data.commu.in$tag_local_identifier), function(x){
 ggplot(aes(x=location_long, y=location_lat, col=ID), data=x)+
  geom_point()#+
  #theme(legend.position = "none") 
})
#remove 37506237_D_2022-02-01

unique(data.commu.out$date)
class(data.commu.out$date)

table(data.commu.out$tag_local_identifier, data.commu.out$date)
table(data.commu.out$tag_local_identifier, data.commu.out$sex)

hist(data.commu.out$ground_speed)
```

### Plot outbound commutes to use

```{r, message=FALSE, warning=FALSE}
# pdf(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/analysis/results/dtw/idDay.pdf")
# lapply(data.commu.out, function(x){
#   ggplot(aes(x=coords.x1.1, y=coords.x2.1, color=as.factor(date)), data=x)+
#     geom_point()+
#     ggtitle(x$tag_local_identifier)
# })
# while (!is.null(dev.list()))  dev.off()

```


### Calculating DTW distance (path similarity) for each individual in February - Colony 1 dry-1

Dynamic Time warping is a method of calculating distance that is more accurate than Euclidean distance. It has an advantage over Euclidean if data points are shifted between each other and we want to look rather at its shape. Additionally two time series don’t have to be equal in length what is an assumption required by the Euclidean distance. The Euclidean distance takes pairs of data points and compares them to each other. DTW calculates the smallest distance between all points - this enables a one-to-many match. In literature dynamic time warping is often paired with k-medoids and hierarchical methods. In some papers it is sometimes paired with k-means though this is a controversial matter (Stack Exchange (2015)). Out of non classical methods, it was also paired with random-swap and hybrid(Aghabozorgi, Shirkhorshidi, and Wah (2015)).

Good explanation of DTW compare to Eucliadina distance: <https://rpubs.com/esobolewska/dtw-time-series>
Example from website:
The main difference between both distances can be best understood graphically. The picture below shows an example of matched points of two data vectors. They were connected based on the minimal distance between points based on DTW (black) and Euclidean (red) distance. It can be seen that with DTW the 11th blue point matches 4 green points. When taking into account the Euclidean distance it can be seen that the assigned points 9-to-9 and 10-to-10 are visibly further than 9-to-11 and 10-to-11. That can significantly impact the overall distance of series between each other. The DTW takes into account the shape of both time series much better.

```{r, echo=FALSE}
# split data by day
feb_id <- split(data.commu.out, f = data.commu.out$tag_local_identifier)

# compute DTW distances within each day between different individuals
results_list <- lapply(names(feb_id), function(id) {
  id_data <- feb_id[[id]]
  
  # split data by day within the individual
  indiv_splits <- split(id_data, id_data$date)
  
  # create all pairwise combinations of individuals for the day
  pairs <- expand.grid(names(indiv_splits), names(indiv_splits))
  pairs <- subset(pairs, Var1 != Var2)
  
# compute DTW distance for each pair of individuals
  dtw_distances <- mapply(function(ind1, ind2) {
    dtw_result <- dtw(indiv_splits[[ind1]][, c("coords.x1", "coords.x2")],
                      indiv_splits[[ind2]][, c("coords.x1", "coords.x2")],
                      keep.internals = TRUE)
    dtw_result$distance  # extracting the DTW distance
  }, pairs$Var1, pairs$Var2, SIMPLIFY = FALSE)
  
  # combine the results with pair information and include the date
  data.frame(id= unique(id_data$tag_local_identifier), date1=pairs$Var1, date2=pairs$Var2, Distance = unlist(dtw_distances))
})

names(results_list) <- names(feb_id)

results_df <- do.call(rbind,results_list)

# add columns of groups id and categorical variable of same group= yes/no
dist_id <- results_df %>%
  mutate(group= str_sub(id, 10,10))

dist_id <- dist_id[!duplicated(dist_id$Distance),]

dist_id_ls <- split(dist_id, dist_id$id) 

# cluster by individual
fitall_ind<- lapply(dist_id_ls, function(x){
  hclust(as.dist(xtabs(Distance ~ date1 + date2, data = x)), method = "ward.D2") #as.dist(xtabs(Distance ~ date1 + date2, data = id))
}) # there is something wrong here

names(fitall_ind) <- names(dist_id_ls)

#remove individuals with only two days of tracking
fitall_ind <- fitall_ind[names(fitall_ind) %in% c("53506935_G", "22507B4E_D", "38500437_D", "38506D37_D")==FALSE]

# plot cluster by id in each day
pdf(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/clustering_byiddate_2022.pdf")
lapply(1:length(fitall_ind), function(i){
  id <- names(fitall_ind[i]) 
  fviz_dend(fitall_ind[[i]], k = 2, color_labels_by_k = FALSE, rect = TRUE, main = id)
})
while (!is.null(dev.list()))  dev.off()

pdf(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/tracksidday.pdf")
lapply(1:length(feb_id), function(i){
  id <- names(feb_id[i]) 
  ggplot(feb_id[[id]], aes(x=location_long, y=location_lat, color=as.factor(date), group =as.factor(date)))+
    geom_path()+
    ggtitle(id)
})
while (!is.null(dev.list()))  dev.off()

# plot one individual
ggplot(aes(x=location_long, y=location_lat, group = date, color=as.factor(date)), data=feb_id[["2C507735_G"]])+
geom_path()+
  facet_wrap(.~date)
```

### Plot some tracks per individual to see how the similar are in their commutes

Inbound commutes look more similar to each other than outbound commutes

```{r}
# look that there is something going on here
outplots <- data.commu.out %>%
  filter( tag_local_identifier== "2C500235_D"| tag_local_identifier== "2C507735_G" |  tag_local_identifier== "37507537_D"| tag_local_identifier== "39506037_G") %>% #tag_local_identifier== "2D507235_G" | |tag_local_identifier=="2E500235_G"
  ggplot(aes(x=location_long, y=location_lat, col=tag_local_identifier, group=ID))+
  scale_color_manual(values=c( "#1F78B4",  "#CAB2D6", "#33A02C", "#B15928"))+  
  geom_path()+
  xlab("longitude")+
  ylab("latitude")+
  theme_classic()+
  theme(legend.position = "none")+
  theme(axis.title = element_text(size=20), axis.title.x = element_text(size = 20), axis.title.y=element_text(size=20), axis.text.y = element_text(size=14), axis.text.x = element_text(size=14))
ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/tracks_out.pdf", width = 5, height = 5)



inplots <- data.commu.in %>%
  filter( tag_local_identifier== "2C500235_D"| tag_local_identifier== "2C507735_G" | tag_local_identifier== "37507537_D"| tag_local_identifier== "39506037_G") %>%#tag_local_identifier== "2D507235_G" | |tag_local_identifier=="2E500235_G"
  ggplot(aes(x=location_long, y=location_lat, col=tag_local_identifier, group=ID))+
  scale_color_manual(values=c( "#1F78B4",  "#CAB2D6", "#33A02C", "#B15928"))+   
  geom_path()+
  xlab("longitude")+
  ylab("latitude")+
  theme_classic()+
  theme(legend.position = "none")+
  theme(axis.title = element_text(size=20), axis.title.x = element_text(size = 20), axis.title.y=element_text(size=20), axis.text.y = element_text(size=14), axis.text.x = element_text(size=14))
ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/tracks_in.pdf", width = 5, height = 5)

```

### Clustering the outbound commutes by ID day for all individuals in February 2022- Colony 1 dry-1

Here we use dtw to calculate distance dissimilarities between commuting track by individual day. We calculate all the pairwise comparison between individual in the different days.
Dynamic Time Warping finds the optimal alignment between the two time series by computing the minimal cumulative distance path through this matrix. The resulting alignment stretches or compresses the sequences to match each other as closely as possible, minimizing the total distance (or "warping cost").


To qualitatively described the commuting paths by individual, I used a clustering method from the hclust function. The clustering method was ward.D2 which calculates minimum variance. It aims to find compact and spherical clusters. This cluster seems to capture better the similarity within individuals compare to the complete linkage which find similar clusters. "ward.D2" implements that criterion (Murtagh and Legendre 2014). With ward.D2, the dissimilarities are squared before cluster updating, compare to ward.D.

```{r}
# split data by day
feb_out <- data.commu.out%>%
  filter(date>="2022-01-31" & date<="2022-02-10")

feb_out_idday <- split(feb_out, f = feb_out$ID)

# function to compute DTW distances between two dataframes
compute_dtw_between_dataframes <- function(df1, df2) {
  dtw_result <- dtw(df1[, c("coords.x2", "coords.x1")],
                    df2[, c("coords.x2", "coords.x1")],
                    keep.internals = TRUE, step.pattern = symmetric2, open.end=TRUE)
  dtw_result$distance  # Extracting the DTW distance
}

# create all pairwise combinations of dataframes
pairs <- expand.grid(names(feb_out_idday), names(feb_out_idday))
pairs <- subset(pairs, Var1 != Var2)

set.seed(1)
# compute DTW distances for each pair of dataframes
dtw_distances <- mapply(function(df1_name, df2_name) {
  df1 <- feb_out_idday[[df1_name]]
  df2 <- feb_out_idday[[df2_name]]
  compute_dtw_between_dataframes(df1, df2)
}, pairs$Var1, pairs$Var2, SIMPLIFY = FALSE)



# combine the results with pair information
all_results <- data.frame(Pair1 = pairs$Var1, Pair2 = pairs$Var2, Distance = unlist(dtw_distances))

# save results
# save(all_results, file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/analysis/data/dtw_feb_bydate.RData")

# add columns of groups id and categorical variable of same group= yes/no
all_results <- all_results %>%
  mutate(id1= str_sub(Pair1, 1,8), id2= str_sub(Pair2, 1,8)) %>%
  mutate(sameid=ifelse(id1 == id2, "yes", "no"), dyad=paste(id1,id2, sep="_"), group1= str_sub(Pair1, 10,10), group2= str_sub(Pair2, 10,10), samegroup=ifelse(group1 == group2, "yes", "no"), group=paste(group1,group2, sep="_"), date1=str_sub(Pair1, 12,21), date2=str_sub(Pair2, 12,21), samedate=ifelse(date1 == date2, "yes", "no"), direc="outbound", season="dry", cave="lagruta")

# change labels of groups
all_results$group[which(all_results$group=="D_G")] <- "G_D"


# Remove duplicates based on 'Distance' column
all_results <- all_results[!duplicated(all_results$Distance), ]

all_results_samedate <- all_results%>%
  filter(sameid=="yes")

#remove individuals with only two days of tracking for the clustering representation
all_results_sub <- all_results %>%
  filter(id1!="53506935" & id1!= "22507B4E" & id1!= "38500437" & id1!="38506D37" & id2!="53506935" & id2!= "22507B4E" & id2!= "38500437" & id2!="38506D37")

all_results_sub$Pair1 <- droplevels(all_results_sub$Pair1)
all_results_sub$Pair2 <- droplevels(all_results_sub$Pair2)
```
### Quickly plot a cluster for all to visualize outbound commutes in February 2022
```{r}
set.seed(1)
fit_idday <- hclust(as.dist(xtabs(Distance ~ Pair1 + Pair2, data = all_results_sub)), method = "ward.D2")

# plot the dendrogram
fviz_dend(fit_idday, k = 15,  color_labels_by_k = TRUE, rect = TRUE,show_labels = TRUE)

ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/cluster_allfeb_idday_ward2_16092024_.pdf", height=20, width =25)
# 
# set.seed(1)
# fit_idday <- hclust(as.dist(xtabs(Distance ~ Pair1 + Pair2, data = all_results)), method = "complete")
# fviz_dend(fit_idday, k=15, color_labels_by_k = TRUE, rect = TRUE,show_labels = TRUE)
# 
# ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/cluster_allfeb_idday_complete19082024_meters.pdf", height=20, width =25)
```
The previous dendogram color by cluster but it is no able to color by individual day even if it falls in a different cluster. I try to do this here in the next figure, to see better where the individuals fall within all the possible clusters inside the tree.

### Cluster analysis for outbound commutes colored by ID in Colony 1 dry-1

```{r}
library(factoextra)
library(dendextend)
library(viridis) 
library(RColorBrewer)

# Example dendrogram data and clustering (you can replace this with your data)
dend <- as.dendrogram(fit_idday) 

# Define a custom color palette or color vector for bars
color_palette <- c("#A6CEE3","#1F78B4", "#B2DF8A", "#33A02C", "#FB9A99", "#E31A1C", "#FDBF6F", "#FF7F00", "#CAB2D6", "#6A3D9A", "#FFFF99", "#B15928")

# match colors with your labels (labels2)
labels1 <- str_sub(fit_idday$labels, 1,10)
labels1_unique <- unique(labels1)
label_colors1 <- setNames(color_palette[1:length(labels1_unique)], labels1_unique)
colored_labels1 <- label_colors1[labels1]
# 
# # Create the fviz_dend plot
# set.seed(1)
# fviz_dend(dend, k = 12, color_labels_by_k = FALSE, rect = FALSE, show_labels = TRUE,
#           label_cols = colored_labels)

# Overlay colored bars (this needs to be done using base R functions after plotting fviz_dend)
# Adjust margins

pdf(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/dtw_febout_ward2.pdf", height=20, width =25)
par(mar = c(4,1,1,12))

# Plot dendrogram again
dend %>% dendextend::set("labels_cex", 1.2) %>% plot(horiz=TRUE)

# Add colored bars
colored_bars(cbind(colored_labels1), dend, horiz = TRUE, rowLabels = "id", cex.rowLabels=2)

# Add legend for the colors
legend("topleft", legend = labels1_unique, fill = color_palette[1:length(labels1_unique)], text.font = 2, cex = 3)
dev.off()

```

### DTW distance for inbound commutes by ID day in Feburary 2022, Colony 1 dry-1

```{r}
# split data by day
feb_in <- data.commu.in%>%
  filter(date>="2022-01-31" & date<="2022-02-10")

feb_in_idday <- split(feb_in, f = feb_in$ID)

# create all pairwise combinations of dataframes
pairs_in_feb <- expand.grid(names(feb_in_idday), names(feb_in_idday))
pairs_in_feb <- subset(pairs_in_feb, Var1 != Var2)

set.seed(1)
# compute DTW distances for each pair of dataframes
dtw_dist_in_feb <- mapply(function(df1_name, df2_name) {
  df1 <- feb_in_idday[[df1_name]]
  df2 <- feb_in_idday[[df2_name]]
  compute_dtw_between_dataframes(df1, df2)
}, pairs_in_feb$Var1, pairs_in_feb$Var2, SIMPLIFY = FALSE)

# combine the results with pair information
result_dist_feb_in <- data.frame(Pair1 = pairs_in_feb$Var1, Pair2 = pairs_in_feb$Var2, Distance = unlist(dtw_dist_in_feb))

# save results
# save(all_results, file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/analysis/data/dtw_feb_bydate.RData")

# add columns of groups id and categorical variable of same group= yes/no
result_dist_feb_in <- result_dist_feb_in %>%
  mutate(id1= str_sub(Pair1, 1,8), id2= str_sub(Pair2, 1,8)) %>%
  mutate(sameid=ifelse(id1 == id2, "yes", "no"), dyad=paste(id1,id2, sep="_"), group1= str_sub(Pair1, 10,10), group2= str_sub(Pair2, 10,10), samegroup=ifelse(group1 == group2, "yes", "no"), group=paste(group1,group2, sep="_"), date1=str_sub(Pair1, 12,21), date2=str_sub(Pair2, 12,21), samedate=ifelse(date1 == date2, "yes", "no"),  direc="inbound", season="dry", cave="lagruta")

# change labels of groups
result_dist_feb_in$group[which(result_dist_feb_in$group=="D_G")] <- "G_D"

# Remove duplicates based on 'time_diff' column
result_dist_feb_in <- result_dist_feb_in[!duplicated(result_dist_feb_in$Distance), ]

#remove individulas with only two days of tracking for the clustering representation
result_dist_feb_in_sub <- result_dist_feb_in %>%
  filter(id1!="53506935" & id1!= "22507B4E" & id1!= "38500437" & id1!="38506D37" & id2!="53506935" & id2!= "22507B4E" & id2!= "38500437" & id2!="38506D37")

result_dist_feb_in_sub$Pair1 <- droplevels(result_dist_feb_in_sub$Pair1)
result_dist_feb_in_sub$Pair2 <- droplevels(result_dist_feb_in_sub$Pair2)
```
### Quickly plot a cluster for all to visualize inbound commute sin February 2022
```{r}
set.seed(1)
fit_idday_in_feb <- hclust(as.dist(xtabs(Distance ~ Pair1 + Pair2, data = result_dist_feb_in_sub)), method = "ward.D2")

# Plot the dendrogram
fviz_dend(fit_idday_in_feb, k = 15,  color_labels_by_k = TRUE, rect = TRUE,show_labels = TRUE, horiz=TRUE, phylo_layout="layout_with_drl")

ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/cluster_feb_in_ward2_trial.pdf", height=20, width =25)
# 
# set.seed(1)
# fit_idday <- hclust(as.dist(xtabs(Distance ~ Pair1 + Pair2, data = all_results)), method = "complete")
# fviz_dend(fit_idday, k=15, color_labels_by_k = TRUE, rect = TRUE,show_labels = TRUE)
# 
# ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/cluster_allfeb_idday_complete19082024_meters.pdf", height=20, width =25)
```
The previous dendogram of inbound commutes colored by cluster but it is no able to color by individual day even if it falls in a different cluster. I try to do this here in the next figure, to see better where the individuals fall within all the possible clusters inside the tree.

### Cluster analysis for inbound commutes colored by ID in Colony 1 dry-1

```{r}
library(factoextra)
library(dendextend)

# Example dendrogram data and clustering (you can replace this with your data)
dend_in_feb <- as.dendrogram(fit_idday_in_feb) 

# Define a custom color palette or color vector for bars
color_palette <- c("#A6CEE3","#1F78B4", "#B2DF8A", "#33A02C", "#FB9A99", "#E31A1C", "#FDBF6F", "#FF7F00", "#CAB2D6", "#6A3D9A", "#FFFF99", "#B15928", "grey")
# match colors with your labels (labels2)
labels2 <- str_sub(fit_idday_in_feb$labels, 1,10)
labels2_unique <- unique(labels2)
label_colors <- setNames(color_palette[1:length(labels2_unique)], labels2_unique)
colored_labels2 <- label_colors[labels2]
# 
# # Create the fviz_dend plot
# set.seed(1)
# fviz_dend(dend, k = 12, color_labels_by_k = FALSE, rect = FALSE, show_labels = TRUE,
#           label_cols = colored_labels)

# Overlay colored bars (this needs to be done using base R functions after plotting fviz_dend)
# Adjust margins

pdf(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/dtw_feb_in_ward2.pdf", height=20, width =20)
par(mar = c(4,1,2,12))

# Plot dendrogram again
dend_in_feb %>% dendextend::set("labels_cex", 1.2) %>% plot(horiz=TRUE)#

# Add colored bars
colored_bars(cbind(colored_labels2), dend_in_feb, horiz = TRUE, rowLabels = "id", cex.rowLabels=2)

# Add legend for the colors
legend("topleft", legend = labels2_unique, fill = color_palette[1:length(labels2_unique)], text.font = 2, cex = 2.5)
dev.off()

pdf(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/dtw_feb_outin.pdf", height=20, width =20)
# Adjust margins and plot the first dendrogram
par(mar = c(4, 1, 1, 12))
dend %>% dendextend::set("labels_cex", 1.2) %>% plot(horiz=TRUE)

# Add colored bars and legend for the first dendrogram
colored_bars(cbind(colored_labels1), dend, horiz = TRUE, rowLabels = "id", cex.rowLabels=2)
legend("topleft", legend = labels1_unique, fill = color_palette[1:length(labels1_unique)], text.font = 2, cex = 3)

# Adjust margins and plot the second dendrogram
#par(mar = c(4, 1, 2, 12))
dend_in_feb %>% dendextend::set("labels_cex", 1.2) %>% plot(horiz=TRUE)#xlim=c(0,max(dist(fit_idday_in_feb$height)))

# Add colored bars and legend for the second dendrogram
colored_bars(cbind(colored_labels2), dend_in_feb, horiz = TRUE, rowLabels = "id", cex.rowLabels=2)
#legend("topleft", legend = labels2_unique, fill = color_palette[1:length(labels2_unique)], text.font = 2, cex = 2.5)
dev.off()
```

### Test if there are more similarities within than between individuals in their paths similarities

```{r}
library(lme4)

# check the distribution of distance
library(fitdistrplus)
fit <- fitdist(all_results$Distance, "weibull")  # Change "weibull" to other distributions
summary(fit)
gofstat(fit)

# diagnostic plots
plot(fit)
cdfcomp(fit)
qqcomp(fit)

# fit multiple distributions
fit_norm <- fitdist(all_results$Distance, "norm")
fit_gamma <- fitdist(all_results$Distance, "gamma")
fit_weib <- fitdist(all_results$Distance, "weibull")

# compare fits visually
# assuming fit_norm has estimated parameters mean (mle) and standard deviation (sd)
plotdist(all_results$Distance, "norm", para = list(mean = fit_norm$estimate["mean"], sd = fit_norm$estimate["sd"]), histo = TRUE, demp = TRUE)
cdfcomp(list(fit_norm, fit_weib), legendtext = c("Normal", "Weibull"))
# weibull is the best distribution

# plot outbound similarity between and within ids
ggplot(aes(x=sameid, y=Distance), data=all_results) +
  geom_jitter(aes(alpha=0.5))+
stat_summary(fun = mean, color = "red") +
  stat_summary(
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "red",
    width = .3
  )

ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/dtw_outbound_feb.pdf")

# plot inbound similarity between and within ids
ggplot(aes(x=sameid, y=Distance), data=result_dist_feb_in) +
  geom_jitter(aes(alpha=0.5))+
stat_summary(fun = mean, color = "red") +
  stat_summary(
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "red",
    width = .3
  )

ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/dtw_inbound_feb.pdf")


# test similarity of outbound commutes in Feburary
all_results$date <- paste(all_results$date1, all_results$date2, sep="_")

m_dtw_out <- lmer(log(Distance)~sameid + (1|dyad) + (1|date), data = all_results)

summary(m_dtw_out)
car::Anova(m_dtw_out)

# plot histogram for distance
hist(log(result_dist_feb_in$Distance))

# test similarity of inbound commutes in Feburary
result_dist_feb_in$date <- paste(result_dist_feb_in$date1, result_dist_feb_in$date2, sep="_")

m_dtw_in <- lmer(log(Distance)~sameid + (1|dyad) + (1|date), data = result_dist_feb_in)

summary(m_dtw_in)
car::Anova(m_dtw_in)
```
Here we see that independent of the outbound and inbound commutes, individuals have more similar paths across the days than path between other individuals in the colony 1 dry-1.

Now we are interested in seeing if outbound commutes are more similar than inbound commutes within the same individuals, to see if there are some particular routes preferred when bats fly to the foraging areas vs. returning to the roost.


### Compare inbound and outbound from the same individual same day in February

```{r}
data.commu.out$direc <- "outbound"
data.commu.in$direc <- "inbound"

febpaths <- rbind(data.commu.out,data.commu.in)

febpaths_id <- split(febpaths, febpaths$tag_local_identifier)

# # compute DTW distances within each day between different individuals
# results_inout <- lapply(names(febpaths_id), function(id) {
#   id_data <- febpaths_id[[id]]
#   
#   # split data by day within the individual
#   indiv_splits <- split(id_data, id_data$direc)
#   
#   # create all pairwise combinations of individuals for the day
#   # pairs <- expand.grid(names(indiv_splits), names(indiv_splits))
#   # pairs <- subset(pairs, Var1 != Var2)
#   
# # compute DTW distance for each pair of individuals
#     dtw_result <- dtw(indiv_splits[[1]][, c("coords.x1", "coords.x2")],
#                       indiv_splits[[2]][, c("coords.x1", "coords.x2")],
#                       keep.internals = TRUE)
#     dist <- dtw_result$distance  # extracting the DTW distance
#   
#   # combine the results with pair information and include the date
#   data.frame(id= unique(id_data$tag_local_identifier), date=unique(id_data$date), Distance = unlist(dist))
# })
# 
# names(results_inout) <- names(febpaths_id)
# 
# results_inout_df <- do.call(rbind, results_inout)
# results_inout_df$season <- "2022_feb"


######## try for all pairwise comparisons


# Compute pairwise DTW distances between outbound and inbound commutes across all date combinations
results_inout <- lapply(names(febpaths_id), function(id) {
  id_data <- febpaths_id[[id]]
  
  # Split data by direction (outbound and inbound)
  direction_splits <- split(id_data, id_data$direc)
  
  # Check if both outbound and inbound data exist
  if (!("outbound" %in% names(direction_splits)) || !("inbound" %in% names(direction_splits))) {
    return(NULL) # Skip if one direction is missing
  }
  
  outbound_data <- direction_splits$outbound
  inbound_data <- direction_splits$inbound
  
  # Create all combinations of dates between outbound and inbound commutes
  date_combinations <- expand.grid(outbound_date = unique(outbound_data$date),
                                    inbound_date = unique(inbound_data$date),
                                    stringsAsFactors = FALSE)
  
  # Calculate DTW distance for each combination
  distances <- apply(date_combinations, 1, function(row) {
    out_date <- row["outbound_date"]
    in_date <- row["inbound_date"]
    
    # Filter data for the specific dates
    out_coords <- outbound_data[outbound_data$date == out_date, c("coords.x1", "coords.x2")]
    in_coords <- inbound_data[inbound_data$date == in_date, c("coords.x1", "coords.x2")]
    
    if (nrow(out_coords) > 0 && nrow(in_coords) > 0) {
      # Compute DTW distance
      dtw_result <- dtw(out_coords, in_coords, keep.internals = TRUE)
      return(dtw_result$distance)
    } else {
      return(NA) # Handle cases where no data is available
    }
  })
  
  # Combine the results
  results <- data.frame(
    id = id,
    outbound_date = date_combinations$outbound_date,
    inbound_date = date_combinations$inbound_date,
    Distance = distances
  )
  
  return(results)
})

# Combine results for all individuals
results_inout_df <- do.call(rbind, results_inout)

# View final results
head(results_inout_df)

results_inout_df$season <- "2022_feb"

# plot boxplot of similarities
ggplot(aes(x=id, y=Distance), data=results_inout_df) +
  geom_boxplot()+
  geom_point()+
stat_summary(fun = mean, color = "red") +
  stat_summary(
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "red",
    width = .3
  )+
  ylab("distance similarity")+
  theme_classic()+
  theme(axis.title = element_text(face = 'bold', size=20), axis.title.x = element_text(size = 20), axis.title.y=element_text(size=20), axis.text.y = element_text(size=14), axis.text.x = element_text(size=14))
ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/inout_sim_byid.pdf")

```

I did DTW distance and clustering for the individuals in February, and now I will do it for the rest of the individuals in March 2022, colony 2, and the individuals in 2023 colony 2

### Individuals on March 2022 - Males in Colony 2

```{r}
# filter to march data
march_males <- bats_commu%>%
  filter(date>="2022-03-01" & date<="2022-03-30")

# split data frame
splitmarch_males <- split(march_males, f=march_males$ID)

# filter individuals with incomplete data
Commu_march_commales <- splitmarch_males[names(splitmarch_males) %in% c("PH_TS_011_2022-03-08","PH_TS_011_2022-03-12", "PH_TS_011_2022-03-15", "PH_TS_011_2022-03-17", "PH_TS_014_2022-03-07", "PH_TS_014_2022-03-08", "PH_TS_014_2022-03-09", "PH_TS_014_2022-03-10", "PH_TS_016_2022-03-09","PH_TS_016_2022-03-10", "PH_TS_080_2022-03-09", "PH_TS_080_2022-03-10", "PH_TS_080_2022-03-11", "PH_TS_080_2022-03-14", "PH_TS_080_2022-03-15", "PH_TS_080_2022-03-16", "PH_TS_080_2022-03-17", "PH_TS_080_2022-03-18", "PH_TS_080_2022-03-19", "PH_TS_080_2022-03-20", "PH_TS_080_2022-03-21", "PH_TS_100_2022-03-07", "PH_TS_100_2022-03-10", "PH_TS_100_2022-03-11", "PH_TS_100_2022-03-16", "PH_TS_100_2022-03-17", "PH_TS_100_2022-03-18", "PH_TS_112_2022-03-08", "PH_TS_112_2022-03-11", "PH_TS_120_2022-03-08") == TRUE]#remove "PH_TS_120_2022-03-08", "PH_TS_120_2022-03-11"

Commu_march_malesc <- as.data.frame(do.call(rbind, Commu_march_commales))

# plot time vs id to see gaps in data
ggplot(aes(x=location_long, y=location_lat, color=as.factor(date)), data=Commu_march_malesc)+
    geom_point()+
    facet_wrap(.~date)

# ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/paths_males_ajcave.pdf", height=20, width =25)

# for loop to divide each day of the animal in commuting outbound and commuting inbound using around 30 minutes as a threshold value
mylist2 <- list() 
for (i in 1:length(Commu_march_commales)){
  x <- which(Commu_march_commales[[i]]$diff>=1215)
  mylist2[[i]] <- split(Commu_march_commales[[i]], cumsum(1:nrow(Commu_march_commales[[i]])%in%x))
  indv <- unique(names(Commu_march_commales[i]))
  names(mylist2[[i]]) <- paste(indv, seq_along(mylist2[[i]]), sep = "_")
}

# convert nested list to a data frame
data_march <- find_df(mylist2)

# add names to list of the commuting outbound or inbound to the list
data_march <- Map(f, data_march, names(data_march))

# convert list to data frame
data.march <- as.data.frame(do.call(rbind, data_march))

data.march <- data.march %>%
  filter(ID!="PH_TS_011_2022-03-15", ID!="PH_TS_011_2022-03-17", ID!="PH_TS_014_2022-03-08", ID!="PH_TS_014_2022-03-07", ID!= "PH_TS_079_2022-03-16", ID!= "PH_TS_100_2022-03-17", ID!= "PH_TS_100_2022-03-18", ID!= "PH_TS_080_2022-03-09")

# remove ID
data.march <- data.march%>%
  filter(name!="PH_TS_016_2022-03-10_1")

# rename some files to make sure classificiation is right
data.march$name[which(data.march$name=="PH_TS_016_2022-03-10_2")] <- "PH_TS_016_2022-03-10_1"
data.march$name[which(data.march$name=="PH_TS_016_2022-03-10_3")] <- "PH_TS_016_2022-03-10_2"

# select only outbound commuting
commu.march.out <- data.march[grep("_1\\b", data.march$name),]#exact match to avoid selecting the inbound commuting tracks
commu.march.out <- commu.march.out%>% 
  filter(ID!="PH_TS_011_2022-03-15", ID!="PH_TS_011_2022-03-17", ID!= "PH_TS_079_2022-03-16", ID!= "PH_TS_100_2022-03-17", ID!= "PH_TS_100_2022-03-18",ID!="PH_TS_120_2022-03-11")

# select inbound commutes
commu.march.in <- data.march[grep("_2\\b", data.march$name),]
commu.march.in <- commu.march.in %>%
  filter( ID!= "PH_TS_079_2022-03-16", ID!= "PH_TS_080_2022-03-09")

# plot outbound commutes
ggplot(commu.march.out, aes(x=location_long, y=location_lat, group=ID, color=ID))+
  geom_path()#+
  #facet_wrap(.~ID)

# plot inbound commutes
ggplot(commu.march.in, aes(x=location_long, y=location_lat, group = ID, color=ID))+
  geom_path()#+
 # facet_wrap(.~ID)
```

### DTW similarity of outbound commutes for males in march 2022, Colony 2

```{r}
# split dataframe by ID
commu.march.out.ls <- split(commu.march.out, commu.march.out$ID)

# create all pairwise combinations of dataframes
pairs_out_march <- expand.grid(names(commu.march.out.ls), names(commu.march.out.ls))
pairs_out_march <- subset(pairs_out_march, Var1 != Var2)

set.seed(1)

# compute DTW distances for each pair of dataframes
dtw_distances_marchout <- mapply(function(df1_name, df2_name) {
  df1 <- commu.march.out.ls[[df1_name]]
  df2 <- commu.march.out.ls[[df2_name]]
  compute_dtw_between_dataframes(df1, df2)
}, pairs_out_march$Var1, pairs_out_march$Var2, SIMPLIFY = FALSE)

# combine the results with pair information
results_march_out <- data.frame(Pair1 = pairs_out_march$Var1, Pair2 = pairs_out_march$Var2, Distance = unlist(dtw_distances_marchout))

# save results
# save(all_results, file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/analysis/data/dtw_feb_bydate.RData")

# add columns of groups id and categorical variable of same group= yes/no
results_march_out<- results_march_out %>%
  mutate(id1= str_sub(Pair1, 1,9), id2= str_sub(Pair2, 1,9)) %>%
  mutate(sameid=ifelse(id1 == id2, "yes", "no"), dyad=paste(id1,id2, sep="_"), date1=str_sub(Pair1, 11,20), date2=str_sub(Pair2, 11,20), samedate=ifelse(date1 == date2, "yes", "no"), direc="outbound", , season="dry", cave="ajcave")

# Remove duplicates based on 'Distance' column
results_march_out <- results_march_out[!duplicated(results_march_out$Distance), ]

march_out_samedate <- results_march_out%>%
  filter(samedate=="yes")

# quickly plot a cluster for all to visualize, but there is not enough information to see a good pattern
set.seed(1)
fit_idday_marchout <- hclust(as.dist(xtabs(Distance ~ Pair1 + Pair2, data = results_march_out)), method = "ward.D2")

# Plot the dendrogram
fviz_dend(fit_idday_marchout, k = 10,  color_labels_by_k = TRUE, rect = TRUE,show_labels = TRUE, cex = 2)

ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/cluster_march_out_ward2.pdf", height=35, width =25)

# boxplot of similarities in outbound commutes in march within and between individuals
ggplot(aes(x=sameid, y=Distance), data=results_march_out) +
  geom_jitter(aes(alpha=0.5))+
stat_summary(fun = mean, color = "red") +
  stat_summary(
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "red",
    width = .3
  )
```


### DTW similarity of inbound commutes for males in march 2022 in colony 2

```{r}
# split dataframe by ID
commu.march.in.ls <- split(commu.march.in, commu.march.in$ID)

# Create all pairwise combinations of dataframes
pairs_in_march <- expand.grid(names(commu.march.in.ls), names(commu.march.in.ls))
pairs_in_march <- subset(pairs_in_march, Var1 != Var2)

set.seed(1)

# compute DTW distances for each pair of dataframes
dtw_distances_marchin <- mapply(function(df1_name, df2_name) {
  df1 <- commu.march.in.ls[[df1_name]]
  df2 <- commu.march.in.ls[[df2_name]]
  compute_dtw_between_dataframes(df1, df2)
}, pairs_in_march$Var1, pairs_in_march$Var2, SIMPLIFY = FALSE)

# combine the results with pair information
results_march_in <- data.frame(Pair1 = pairs_in_march$Var1, Pair2 = pairs_in_march$Var2, Distance = unlist(dtw_distances_marchin))

# save results
# save(all_results, file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/analysis/data/dtw_feb_bydate.RData")

# add columns of groups id and categorical variable of same group= yes/no
results_march_in<- results_march_in %>%
  mutate(id1= str_sub(Pair1, 1,9), id2= str_sub(Pair2, 1,9)) %>%
  mutate(sameid=ifelse(id1 == id2, "yes", "no"), dyad=paste(id1,id2, sep="_"), date1=str_sub(Pair1, 11,20), date2=str_sub(Pair2, 11,20), samedate=ifelse(date1 == date2, "yes", "no"), direc="inbound", , season="dry", cave="ajcave")

# Remove duplicates based on 'time_diff' column
results_march_in <- results_march_in[!duplicated(results_march_in$Distance), ]


march_in_samedate <- results_march_in%>%
  filter(samedate=="yes")

# quickly plot a cluster for all to visualize
set.seed(1)
fit_idday_marchin <- hclust(as.dist(xtabs(Distance ~ Pair1 + Pair2, data = results_march_in)), method = "ward.D2")

# Plot the dendrogram
fviz_dend(fit_idday_marchin, k = 10,  color_labels_by_k = TRUE, rect = TRUE,show_labels = TRUE, cex = 2)

ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/cluster_march_in_ward2.pdf", height=35, width =25)

#still test this
# boxplot of similarities in outbound commutes in march
ggplot(aes(x=sameid, y=Distance), data=results_march_in) +
  geom_jitter(aes(alpha=0.5))+
stat_summary(fun = mean, color = "red") +
  stat_summary(
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "red",
    width = .3
  )
```


### Compare inbound and outbound from the same individual same day in March

```{r}
commu.march.out$direc <- "outbound"
commu.march.in$direc <- "inbound"

marchpaths <- rbind(commu.march.out,commu.march.in)

marchpaths_id <- split(marchpaths, marchpaths$tag_local_identifier)

# Compute pairwise DTW distances between outbound and inbound commutes across all date combinations
results_inout_march <- lapply(names(marchpaths_id), function(id) {
  id_data <- marchpaths_id[[id]]
  
  # Split data by direction (outbound and inbound)
  direction_splits <- split(id_data, id_data$direc)
  
  # Check if both outbound and inbound data exist
  if (!("outbound" %in% names(direction_splits)) || !("inbound" %in% names(direction_splits))) {
    return(NULL) # Skip if one direction is missing
  }
  
  outbound_data <- direction_splits$outbound
  inbound_data <- direction_splits$inbound
  
  # Create all combinations of dates between outbound and inbound commutes
  date_combinations <- expand.grid(outbound_date = unique(outbound_data$date),
                                    inbound_date = unique(inbound_data$date),
                                    stringsAsFactors = FALSE)
  
  # Calculate DTW distance for each combination
  distances <- apply(date_combinations, 1, function(row) {
    out_date <- row["outbound_date"]
    in_date <- row["inbound_date"]
    
    # Filter data for the specific dates
    out_coords <- outbound_data[outbound_data$date == out_date, c("coords.x1", "coords.x2")]
    in_coords <- inbound_data[inbound_data$date == in_date, c("coords.x1", "coords.x2")]
    
    if (nrow(out_coords) > 0 && nrow(in_coords) > 0) {
      # Compute DTW distance
      dtw_result <- dtw(out_coords, in_coords, keep.internals = TRUE)
      return(dtw_result$distance)
    } else {
      return(NA) # Handle cases where no data is available
    }
  })
  
  # Combine the results
  results <- data.frame(
    id = id,
    outbound_date = date_combinations$outbound_date,
    inbound_date = date_combinations$inbound_date,
    Distance = distances
  )
  
  return(results)
})

# Combine results for all individuals
results_inout_march_df <- do.call(rbind, results_inout_march)

# View final results
head(results_inout_march_df)

# compute DTW distances within each day between different individuals
# results_inout_march <- lapply(names(marchpaths_id), function(id) {
#   id_data <- marchpaths_id[[id]]
#   
#   # split data by day within the individual
#   indiv_splits <- split(id_data, id_data$direc)
#   
#   # create all pairwise combinations of individuals for the day
#   # pairs <- expand.grid(names(indiv_splits), names(indiv_splits))
#   # pairs <- subset(pairs, Var1 != Var2)
#   
# # compute DTW distance for each pair of individuals
#     dtw_result <- dtw(indiv_splits[[1]][, c("coords.x1", "coords.x2")],
#                       indiv_splits[[2]][, c("coords.x1", "coords.x2")],
#                       keep.internals = TRUE)
#     dist <- dtw_result$distance  # extracting the DTW distance
#   
#   # combine the results with pair information and include the date
#   data.frame(id= unique(id_data$tag_local_identifier), date=unique(id_data$date), Distance = unlist(dist))
# })

# names(results_inout_march) <- names(marchpaths_id)

# results_inout_march_df <- do.call(rbind, results_inout_march)
results_inout_march_df$season <- "2022_march"

# plot boxplot of similarities
ggplot(aes(x=id, y=Distance), data=results_inout_march_df) +
  geom_boxplot()+
  geom_point()+
stat_summary(fun = mean, color = "red") +
  stat_summary(
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "red",
    width = .3
  )+
  ylab("distance similarity")+
  theme_classic()
ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/inout_sim_byid_march.pdf", height=10, width =15)

#plot commuting paths
ggplot(marchpaths[marchpaths$tag_local_identifier=="PH_TS_016",], aes(x=location_long, y=location_lat, color=tag_local_identifier, linetype =direc, sgroup=ID))+
  geom_path()+
  theme_classic()

```

### Filter wet season commuting paths in 2023

Here I filtered wet season commuting paths which are complete in order to calculate DTW path similarities

```{r}
wet_bats <- bocas%>%
  filter(behav=="commuting", n>=5)%>%
  filter(date>="2022-03-30")

# add time difference to split commuting tracks by the ID_day
wet_bats <- wet_bats %>%
  group_by(ID_batday) %>%
  mutate(diff = timestamps - lag(timestamps))
wet_bats$diff <- as.numeric(wet_bats$diff)*60
wet_bats$diff[is.na(wet_bats$diff)] <- 0

# split by ID
splitwet <- split(wet_bats, f=wet_bats$ID)

#complete tracks
commu_wet_complete <- splitwet[names(splitwet) %in% c("PHYL11_2023-08-15", "PHYL11_2023-08-16", "PHYL11_2023-08-17", "PHYL11_2023-08-18", "PHYL16_2023-08-14", "PHYL16_2023-08-15",  "PHYL21_2023-08-18", "PHYL25_2023-08-14", "PHYL25_2023-08-15", "PHYL25_2023-08-16", "PHYL25_2023-08-17", "PHYL25_2023-08-18", "PHYL28_2023-08-14", "PHYL28_2023-08-15", "PHYL34_2023-08-13", "PHYL34_2023-08-14",  "PHYL4_2023-08-14", "PHYL7_2023-08-16", "PHYL7_2023-08-17", "PHYL7_2023-08-18","PHYL9_2023-08-14",    "PHYL9_2023-08-18") == TRUE]

# for loop to divide each day of the animal in commuting outbound and commuting inbound using around 30 minutes as a threshold value
mylist3 <- list() 
for (i in 1:length(commu_wet_complete)){
  x <- which(commu_wet_complete[[i]]$diff>=1900)
  mylist3[[i]] <- split(commu_wet_complete[[i]], cumsum(1:nrow(commu_wet_complete[[i]])%in%x))
  indv <- unique(names(commu_wet_complete[i]))
  names(mylist3[[i]]) <- paste(indv, seq_along(mylist3[[i]]), sep = "_")
}

# convert nested list to a data frame
data_wet <- find_df(mylist3)

# add names to list of the commuting outbound or inbound to the list
data_wet <- Map(f, data_wet, names(data_wet))

# convert list to data frame
data.commu.wet <- as.data.frame(do.call(rbind, data_wet))

# filter ID incomplete commutes
data.commu.wet <- data.commu.wet %>%
  filter(ID!="PHYL28_2023-08-14", ID!="PHYL28_2023-08-15", ID!="PHYL34_2023-08-13", ID!="PHYL34_2023-08-14", ID!="PHYL9_2023-08-14", ID!="PHYL9_2023-08-18", ID!="PHYL4_2023-08-14", ID!="PHYL1_2023-08-14", ID!="PHYL25_2023-08-17", ID!="PHYL9_2023-08-18" ,ID!="PHYL7_2023-08-18")

# select only outbound commuting
data.commu.wet.out <- data.commu.wet[grep("_1\\b", data.commu.wet$name),]#exact match to avoid selecting the inbound commuting tracks

# plot outbound commutes to see paths
ggplot(data.commu.wet.out, aes(x=location_long, y=location_lat, group = ID, color=ID))+
  geom_path()+
  facet_wrap(.~tag_local_identifier)

# select only inbound commuting
data.commu.wet.in <- data.commu.wet[grep("_2\\b", data.commu.wet$name),]
data.commu.wet.in <- data.commu.wet.in%>%
  filter()

# plot inbound commutes to see paths
ggplot(data.commu.wet.in, aes(x=location_long, y=location_lat, group = ID, color=ID))+
  geom_path()+
  facet_wrap(.~tag_local_identifier)

# mapview::mapView(data.commu.wet.out, xcol="location_long", ycol="location_lat", zcol="ID", legend=F, crs="EPSG:4326")
# 
# 
# mapview::mapView(data.commu.wet.in, xcol="location_long", ycol="location_lat", zcol="ID", legend=F, crs="EPSG:4326")

```

### DTW for outbound commutes paths in the wet season of 2023

```{r}
#split commuting outbound paths
commu.wet.out.ls <- split(data.commu.wet.out, data.commu.wet.out$ID)

# Create all pairwise combinations of dataframes
pairs_wet_out <- expand.grid(names(commu.wet.out.ls), names(commu.wet.out.ls))
pairs_wet_out <- subset(pairs_wet_out, Var1 != Var2)

set.seed(1)
# Compute DTW distances for each pair of dataframes
distances_wet_out <- mapply(function(df1_name, df2_name) {
  df1 <- commu.wet.out.ls[[df1_name]]
  df2 <- commu.wet.out.ls[[df2_name]]
  compute_dtw_between_dataframes(df1, df2)
}, pairs_wet_out$Var1, pairs_wet_out$Var2, SIMPLIFY = FALSE)

# Combine the results with pair information
results_wet_out <- data.frame(Pair1 = pairs_wet_out$Var1, Pair2 = pairs_wet_out$Var2, Distance = unlist(distances_wet_out))

# save results
# save(all_results, file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/analysis/data/dtw_feb_bydate.RData")

# add columns of groups id and categorical variable of same group= yes/no
results_wet_out<- results_wet_out %>%
  mutate(id1= str_sub(Pair1, end=-12), id2= str_sub(Pair2, end=-12)) %>%
  mutate(sameid=ifelse(id1 == id2, "yes", "no"), dyad=paste(id1,id2, sep="_"), date1=str_sub(Pair1, start=-10), date2=str_sub(Pair2, start=-10), samedate=ifelse(date1 == date2, "yes", "no"), direc="outbound", season="wet")

results_wet_out$cave1 <- data.commu.wet.out$cave[match(results_wet_out$id1, data.commu.wet.out$tag_local_identifier,)]
results_wet_out$cave2 <-data.commu.wet.out$cave[match(results_wet_out$id2, data.commu.wet.out$tag_local_identifier,)]

results_wet_out$samecave <- ifelse(results_wet_out$cave1 == results_wet_out$cave2, "yes", "no")

# Remove duplicates based on 'time_diff' column
results_wet_out <- results_wet_out[!duplicated(results_wet_out$Distance), ]

results_wet_out_samecave <- results_wet_out%>%
  filter(samecave=="yes")

# quickly plot a cluster for all to visualize
set.seed(1)
fit_idday_wet_out <- hclust(as.dist(xtabs(Distance ~ Pair1 + Pair2, data = results_wet_out_samecave)), method = "ward.D2")

# Plot the dendrogram
fviz_dend(fit_idday_wet_out, k =5,  color_labels_by_k = TRUE, rect = TRUE,show_labels = TRUE)

ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/cluster_wet_out_ward2.pdf", height=35, width =25)


# plot boxplot of similarities
ggplot(aes(x=sameid, y=Distance), data=results_wet_out_samecave) +
  geom_jitter(aes(alpha=0.5))+
stat_summary(fun = mean, color = "red") +
  stat_summary(
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "red",
    width = .3
  )

```

### DTW for inbound commutes paths in the wet season of 2023

```{r}
#split commuting outbound paths
commu.wet.in.ls <- split(data.commu.wet.in, data.commu.wet.in$ID)

# Create all pairwise combinations of dataframes
pairs_wet_in <- expand.grid(names(commu.wet.in.ls), names(commu.wet.in.ls))
pairs_wet_in <- subset(pairs_wet_in, Var1 != Var2)

set.seed(1)
# Compute DTW distances for each pair of dataframes
distances_wet_in <- mapply(function(df1_name, df2_name) {
  df1 <- commu.wet.in.ls[[df1_name]]
  df2 <- commu.wet.in.ls[[df2_name]]
  compute_dtw_between_dataframes(df1, df2)
}, pairs_wet_in$Var1, pairs_wet_in$Var2, SIMPLIFY = FALSE)

# Combine the results with pair information
results_wet_in <- data.frame(Pair1 = pairs_wet_in$Var1, Pair2 = pairs_wet_in$Var2, Distance = unlist(distances_wet_in))

# save results
# save(all_results, file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/analysis/data/dtw_feb_bydate.RData")

# add columns of groups id and categorical variable of same group= yes/no
results_wet_in<- results_wet_in %>%
  mutate(id1= str_sub(Pair1, end=-12), id2= str_sub(Pair2, end=-12)) %>%
  mutate(sameid=ifelse(id1 == id2, "yes", "no"), dyad=paste(id1,id2, sep="_"), date1=str_sub(Pair1, start=-10), date2=str_sub(Pair2, start=-10), samedate=ifelse(date1 == date2, "yes", "no"), direc="inbound", season="wet")

results_wet_in$cave1 <- data.commu.wet.in$cave[match(results_wet_in$id1, data.commu.wet.in$tag_local_identifier,)]
results_wet_in$cave2 <-data.commu.wet.in$cave[match(results_wet_in$id2, data.commu.wet.in$tag_local_identifier,)]
results_wet_in$cave <- ifelse(results_wet_in$cave1==results_wet_in$cave2, results_wet_in$cave1, "mixed")

results_wet_in$samecave <- ifelse(results_wet_in$cave1 == results_wet_in$cave2, "yes", "no")

# Remove duplicates based on 'Distance' column
results_wet_in <- results_wet_in[!duplicated(results_wet_in$Distance), ]

results_wet_in_samecave <- results_wet_in%>%
  filter(samecave=="yes")

# quickly plot a cluster for all to visualize
set.seed(1)
fit_idday_in_out <- hclust(as.dist(xtabs(Distance ~ Pair1 + Pair2, data = results_wet_in_samecave)), method = "ward.D2")

# Plot the dendrogram
fviz_dend(fit_idday_in_out, k =5,  color_labels_by_k = TRUE, rect = TRUE,show_labels = TRUE)

ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/cluster_wet_in_ward2.pdf", height=35, width =25)


# plot boxplot of similarities
ggplot(aes(x=sameid, y=Distance), data=results_wet_in_samecave) +
  geom_jitter(aes(alpha=0.5))+
stat_summary(fun = mean, color = "red") +
  stat_summary(
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "red",
    width = .3
  )

```

### Compare inbound and outbound from the same individual same day in Wet season 2023

```{r}
data.commu.wet.out$direc <- "outbound"
data.commu.wet.in$direc <- "inbound"

wetpaths <- rbind(data.commu.wet.out,data.commu.wet.in)

wetpaths_id <- split(wetpaths, wetpaths$tag_local_identifier)

# Compute pairwise DTW distances between outbound and inbound commutes across all date combinations
results_inout_wet <- lapply(names(wetpaths_id), function(id) {
  id_data <- wetpaths_id[[id]]
  
  # Split data by direction (outbound and inbound)
  direction_splits <- split(id_data, id_data$direc)
  
  # Check if both outbound and inbound data exist
  if (!("outbound" %in% names(direction_splits)) || !("inbound" %in% names(direction_splits))) {
    return(NULL) # Skip if one direction is missing
  }
  
  outbound_data <- direction_splits$outbound
  inbound_data <- direction_splits$inbound
  
  # Create all combinations of dates between outbound and inbound commutes
  date_combinations <- expand.grid(outbound_date = unique(outbound_data$date),
                                    inbound_date = unique(inbound_data$date),
                                    stringsAsFactors = FALSE)
  
  # Calculate DTW distance for each combination
  distances <- apply(date_combinations, 1, function(row) {
    out_date <- row["outbound_date"]
    in_date <- row["inbound_date"]
    
    # Filter data for the specific dates
    out_coords <- outbound_data[outbound_data$date == out_date, c("coords.x1", "coords.x2")]
    in_coords <- inbound_data[inbound_data$date == in_date, c("coords.x1", "coords.x2")]
    
    if (nrow(out_coords) > 0 && nrow(in_coords) > 0) {
      # Compute DTW distance
      dtw_result <- dtw(out_coords, in_coords, keep.internals = TRUE)
      return(dtw_result$distance)
    } else {
      return(NA) # Handle cases where no data is available
    }
  })
  
  # Combine the results
  results <- data.frame(
    id = id,
    outbound_date = date_combinations$outbound_date,
    inbound_date = date_combinations$inbound_date,
    Distance = distances
  )
  
  return(results)
})

# Combine results for all individuals
results_inout_wet_df <- do.call(rbind, results_inout_wet)

# compute DTW distances within each day between different individuals
# results_inout_wet <- lapply(names(wetpaths_id), function(id) {
#   id_data <- wetpaths_id[[id]]
#   
#   # split data by day within the individual
#   indiv_splits <- split(id_data, id_data$direc)
#   
#   # create all pairwise combinations of individuals for the day
#   # pairs <- expand.grid(names(indiv_splits), names(indiv_splits))
#   # pairs <- subset(pairs, Var1 != Var2)
#   
# # compute DTW distance for each pair of individuals
#     dtw_result <- dtw(indiv_splits[[1]][, c("coords.x1", "coords.x2")],
#                       indiv_splits[[2]][, c("coords.x1", "coords.x2")],
#                       keep.internals = TRUE)
#     dist <- dtw_result$distance  # extracting the DTW distance
#   
#   # combine the results with pair information and include the date
#   data.frame(id= unique(id_data$tag_local_identifier), date=unique(id_data$date), Distance = unlist(dist))
# })

# names(results_inout_wet) <- names(wetpaths_id)

# results_inout_wet_df <- do.call(rbind, results_inout_wet)
results_inout_wet_df$season <- "2023_august"

# plot boxplot of similarities
ggplot(aes(x=id, y=Distance), data=results_inout_wet_df) +
  geom_boxplot()+
  geom_point()+
stat_summary(fun = mean, color = "red") +
  stat_summary(
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "red",
    width = .3
  )+
  ylab("distance similarity")+
  theme_classic()
# ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/inout_sim_byid_2023.pdf", height=10, width =15)

# outbound and dinbound commutes
ggplot(wetpaths, aes(x=location_long, y=location_lat, color=tag_local_identifier, linetype =direc, sgroup=ID))+
  geom_path()+
  theme_classic() 
```
### Save dataframe with all outbound and inbound commutes
```{r}
allpaths <- rbind(febpaths,marchpaths, wetpaths)

#look at sex by id
table(allpaths$tag_local_identifier, allpaths$sex)

# write.csv(allpaths,file = "/Users/ccalderon/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/data/allout_inbound_paths.csv")
write.csv(allpaths,file = "/Users/ccalderon/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/data/allpairwise_out_inbound_paths.csv")

allpaths <- read.csv(file = "/Users/ccalderon/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/data/allpairwise_out_inbound_paths.csv")
```

### Comparison of out- and inbound distance across all tracking periods

```{r}
#install.packages("gghighlight")
library(gghighlight)
inout_all <- rbind(results_inout_df, results_inout_march_df, results_inout_wet_df)

# add sex
inout_all$sex <- bocas$sex[match(inout_all$id, bocas$tag_local_identifier)]
inout_all$cave_year <- bocas$year_cave[match(inout_all$id, bocas$tag_local_identifier)]
inout_all$samedate <- ifelse(inout_all$outbound_date==inout_all$inbound_date, "yes","no")

ids <- c("37507537_D","39506037_G","PH_TS_080", "PH_TS_100", "PHYL16", "PHYL25")


inout_all_mean <- inout_all %>%
  group_by(cave_year)%>%
  dplyr::summarise(mean_dist=mean(Distance), sd_dist=sd(Distance))

#plot distances
outin_id <- inout_all%>%
  filter(samedate=="yes")%>%
 ggplot(aes(x=season, y=Distance)) +
  geom_boxplot()+#"#A6D854", "#4DAF4A"
  geom_jitter(aes(alpha=0.8, shape = sex), position = position_jitterdodge(jitter.width = 0.4))+
   geom_point(aes(shape = sex), data = inout_all %>% filter(samedate=="yes", id == "37507537_D"), color = "#A6D854", position = position_jitter(width = 0.4), size=3) +
    geom_point(aes(shape = sex),data = inout_all %>% filter(samedate=="yes", id == "39506037_G"), color = "#4DAF4A", position = position_jitter(width = 0.4), size=3) +
    geom_point(aes(shape = sex),data = inout_all %>% filter(samedate=="yes", id == "PH_TS_080"), color = "#756BB1", position = position_jitter(width = 0.4), size=3) +
   geom_point(aes(shape = sex),data = inout_all %>% filter(samedate=="yes", id == "PH_TS_100"), color = "#54278F", position = position_jitter(width = 0.4), size=3) +
  geom_point(aes(shape = sex),data = inout_all %>% filter(samedate=="yes", id == "PHYL11"), color = "#E6AB02", position = position_jitter(width = 0.4), size=3) +
  geom_point(aes(shape = sex),data = inout_all %>% filter(samedate=="yes", id == "PHYL25"), color = "#D95F02", , position = position_jitter(width = 0.4), size=3) +
  #"#CAB2D6","#B15928"
  scale_x_discrete(labels = c("dry 2022-1", "dry 2022-2", "wet 2023"))+#"dry 2022-1", "dry 2022-2", "wet 2023"
#facet_wrap(.~ season)+
stat_summary(fun = mean, color = "red") +
  stat_summary(
    fun.min = function(x) mean(x) - sd(x),
    fun.max = function(x) mean(x) + sd(x),
    geom = "errorbar",
    color = "red",
    width = .3
  )+
  #scale_color_manual(values =c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00"))+
  #gghighlight(id== ids, label_key=ids)+
  ylab("DTW distance between \n out / inbound commutes \n per id")+
  xlab("Period")+
  theme_classic()+
  theme(legend.position = "none", axis.title = element_text(size=20), axis.title.x = element_text(size = 20), axis.title.y=element_text(size=20), axis.text.y = element_text(size=14), axis.text.x = element_text(size=14))

outin_id

ggsave(file = "~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/inout_across_seasons.pdf")
```

### Plot inbound and outbound commute from the same individuals in different nights

```{r}
library(patchwork)

# Define common limits for longitude and latitude
x_limits <- c(-82.50, -82.28)  # Adjust as needed
y_limits <- c(9.37, 9.44)    # Adjust as needed
y_breaks <- seq(9.37, 9.44, by = 0.02)

#display.brewer.pal(n = 8, name = 'Dark2')
p1 <- febpaths %>%
  filter(tag_local_identifier=="39506037_G" | tag_local_identifier=="37507537_D") %>%
  ggplot(aes(x=location_long, y=location_lat, color=tag_local_identifier, linetype =direc, sgroup=ID))+
  scale_color_manual(values =c("#A6D854", "#4DAF4A"))+ #"#E69F00","#56B4E9"
  geom_path(size=1)+
  xlab("Longitude")+
  ylab("Latitude")+
  scale_x_continuous(limits = x_limits) +
  scale_y_continuous(limits = y_limits, breaks = y_breaks) +
  theme_classic()+
  theme(legend.position = "none", axis.title = element_text(size=20), axis.title.x = element_text(size = 20), axis.title.y=element_text(size=20), axis.text.y = element_text(size=14), axis.text.x = element_text(size=14))


p2 <- marchpaths %>%
  filter(tag_local_identifier=="PH_TS_080" | tag_local_identifier=="PH_TS_014") %>%
  ggplot(aes(x=location_long, y=location_lat, color=tag_local_identifier, linetype =direc, sgroup=ID))+
  scale_color_manual(values =c("#756BB1", "#54278F"))+ #"#009E73","#F0E442"
  geom_path(size=1)+
  xlab("Longitude")+
  ylab("Latitude")+
  scale_x_continuous(limits = x_limits) +
  scale_y_continuous(limits = y_limits, breaks = y_breaks) +
  theme_classic()+
  theme(legend.position = "none", axis.title = element_text(size=20), axis.title.x = element_text(size = 20), axis.title.y=element_text(size=20), axis.text.y = element_text(size=14), axis.text.x = element_text(size=14))



p3 <- wetpaths %>%
  filter(tag_local_identifier=="PHYL25" | tag_local_identifier=="PHYL7") %>%
  ggplot(aes(x=location_long, y=location_lat, color=tag_local_identifier, linetype =direc, sgroup=ID))+
  scale_color_manual(values =c("#E6AB02", "#D95F02"))+ #"#0072B2","#D55E00"
  geom_path(size=1)+
  xlab("Longitude")+
  ylab("Latitude")+
  scale_x_continuous(limits = x_limits) +
  scale_y_continuous(limits = y_limits, breaks = y_breaks) +
  theme_classic()+
  theme(legend.position = "none", axis.title = element_text(size=20), axis.title.x = element_text(size = 20), axis.title.y=element_text(size=20), axis.text.y = element_text(size=14), axis.text.x = element_text(size=14))


tracks <- p1/p2/p3/ plot_layout(axes = "collect")

tracks
```


### Does outbound and inbound commute distances are more different than the distances betweeen outbound-outbound and inbound-inbound commutes?
```{r}
library(lme4)
distsim_all <- all_results %>%
  bind_rows(result_dist_feb_in, results_march_out, results_march_in,results_wet_out,results_wet_in) %>%
  dplyr::select("Pair1", "Pair2", "Distance" , "id1", "id2", "sameid", "dyad", "date1", "date2", "samedate", "direc",  "season", "cave")

distsim_all$year_cave <- "dry 2022-1"

distsim_all$year_cave[which(distsim_all$season=="dry" & distsim_all$cave=="ajcave")] <- "dry 2022-2"

distsim_all$year_cave[which(distsim_all$season=="wet")] <- "wet 2023"

write.csv(distsim_all, file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/data/distsimtable_commutes.csv")
#distsim_all <- read.csv(file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/data/distsimtable_commutes.csv")

distsim_all$sameid <- factor(distsim_all$sameid, levels=c("yes","no"))

#inout_all$day_year <- yday(inout_all$date)
inout_all$direc <- "both"
inout_all$cave_year[inout_all$cave_year=="2022_lagruta_Feb"] <- "dry 2022-1"
inout_all$cave_year[inout_all$cave_year=="2022_ajcave"] <- "dry 2022-2"
inout_all$cave_year[inout_all$cave_year=="2023_ajcave"] <- "wet 2023"
colnames(inout_all)[7] <- "year_cave"

#filter only same id data
distsim_all_sameid <- distsim_all%>%
  filter(sameid=="yes")

distsim_all_sameid$date <- paste(distsim_all_sameid$date1, distsim_all_sameid$date2, sep="_")
distsim_all_sameid$id <- str_sub(distsim_all_sameid$Pair1, end = -12)
distsim_all_sameid$sex <- inout_all$sex[match(distsim_all_sameid$id,inout_all$id)]


# merge distance from same id to the outbound and inbound distance difference by individual by night
inout_all <- rbind(inout_all[,c("id", "Distance", "direc", "year_cave", "sex")], distsim_all_sameid[,c("id", "Distance", "direc", "year_cave", "sex")])


m_diff_inout <- lmer(log(Distance)~year_cave*direc +(1|id), data=inout_all)

summary(m_diff_inout)
car::Anova(m_diff_inout)
library(MuMIn)
r2 <- r.squaredGLMM(m_diff_inout)
print(r2)

plot(m_diff_inout)

acf(residuals(m_diff_inout))

# Marginal means for the interaction of sameid and season
library(emmeans)
emm_interaction <- emmeans(m_diff_inout, ~ direc*year_cave)

# Pairwise comparisons for the interaction
pairwise_interaction <- pairs(emm_interaction, adjust = "tukey")
pairwise_interaction

mean_dist_inout <- inout_all%>%
  group_by(year_cave, direc)%>%
  dplyr::summarise(meandist=mean(Distance, na.rm = TRUE), sddist=sd(Distance, na.rm = TRUE))
```


### Plot similarity within and between individuals of the outbound and inbound commutes

```{r}
library(ggh4x)# for dynamic color in face wraps

# only color strips in x-direction
strip <- strip_themed(background_x = elem_list_rect(fill = c("#756BB1", "#21918c","#F0E442"), alpha=c(0.5)))

# Necessary to put RH% into the facet labels
period_names <- as_labeller(
     c(`dry 2022-1` = "Dry 2022-1", `dry 2022-2` = "Dry 2022-2",`wet 2023` = "Wet 2023"))
### plot path similarity inbound and outbound commutes
dtw_seasons <- ggplot(aes(x=direc,y = Distance, color=sameid, group = sameid), data=distsim_all) +
  #geom_boxplot(position=position_dodge(1))+
  geom_jitter(aes(y = Distance, alpha=sameid), position = position_jitterdodge(jitter.width = 0.3, jitter.height = 0.05))+
  #scale_color_manual(values = c("#440154", "#21918c", "#fde725"))+
  scale_alpha_manual(values = c(0.8,0.1))+
  scale_color_manual(values = c("black","grey"))+
  guides(color="none")+
  #scale_alpha_manual(values=c(0.1,0.8))+
  facet_wrap2(.~year_cave, strip=strip, labeller = period_names)+
  stat_summary(aes(y = Distance, group = sameid), fun = mean, 
               color="red",
               position = position_dodge(.8)) +#FF9999
  stat_summary(aes(y = Distance, group = sameid), 
               fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    #color="black",
    color="red",
    width = .3, 
    position = position_dodge(.8)
  ) +
  ylab("DTW pairwise distances")+
  xlab("Path direction")+
  theme_classic()+ 
  theme(legend.position = "none", axis.title = element_text(size=20), axis.title.x = element_text(size = 20), axis.title.y=element_text(size=20), axis.text.y = element_text(size=14), axis.text.x = element_text(size=14), strip.text.x = element_text(size = 12, color="black"))

dtw_seasons

ggsave(file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/inoutflight_simi_plot3_colors.pdf", width=10, height = 5)


dtw_seasons / (outin_id + tracks)+ plot_annotation(tag_levels = "A") &
  theme(
    plot.tag = element_text(size = 20)  # Customize tag size and appearance
  )
  

ggsave(file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper3/analysis/results/dtw/DTW_joined.pdf", width=12, height = 12)
```

### Plot outbound commutes by individual 

```{r}
# plot outbound similarity between and within ids
distsim_all %>% 
  filter(sameid=="yes") %>% 
  #filter(id1!="53506935", id1!= "22507B4E", id1!="38500437", id1!="38506D37") %>%
  ggplot(aes(x=id1, y=Distance, color=direc)) +
  #geom_jitter(aes(alpha=0.5))+
stat_summary(aes(y = Distance, group = direc), fun = mean,
    position = position_dodge(.8)) +
  stat_summary(aes(y = Distance, group = direc),
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "black",
    width = .3, 
    position = position_dodge(.8)
  )+ theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
axis.title.x = element_blank())#annotate("text",x = 30, y = 12, label = "R = 0.22*", size = 4

# plot outbound and commutes by period  
distsim_all %>% 
  filter(sameid=="no") %>% 
  #filter(id1!="53506935", id1!= "22507B4E", id1!="38500437", id1!="38506D37") %>%
  ggplot(aes(x=year_cave, y=Distance, color=direc)) +
  #geom_jitter(aes(alpha=0.5))+
stat_summary(aes(y = Distance, group = direc), fun = mean,
    position = position_dodge(.8)) +
  stat_summary(aes(y = Distance, group = direc),
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "black",
    width = .3, 
    position = position_dodge(.8)
  )+ theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
axis.title.x = element_blank())
```

### Calculate repeatability for individuals in February for outbound commutes with Rptr
```{r}
library(rptR)

outbound <- distsim_all %>%
  filter(direc=="outbound", sameid=="yes", year_cave=="dry 2022-1")%>%
  mutate(date=paste(date1, date2, sep="_"))

outbound$day_year1 <- yday(outbound$date1)
outbound$day_year2 <- yday(outbound$date2)
outbound$time_diff <- abs(outbound$day_year1-outbound$day_year2)


outbound$cave[which(is.na(outbound$cave))] <- "ajcave_2023"
# Check the levels and counts
table(outbound$cave)

# Check for NAs in Distance, date, and id1
sum(is.na(outbound$Distance))
sum(is.na(outbound$date))
sum(is.na(outbound$id1))

# Convert id1 to factor if necessary
outbound$id1 <- as.factor(outbound$id1)

###Repeatability of the model
#split by cave and calculate repeatability by population
out_rep<- rpt(log(Distance)~time_diff + (1|id1), grname = c("id1","Residual","Fixed"), 
    data = outbound, datatype = "Gaussian", nboot = 1000, npermut = 0, ratio = FALSE, adjusted = FALSE)

print(out_rep)
0.102/(0.184 +0.02)
0.045/(0.011 +0.002)

plot(out_rep, grname = "id1", scale = "probit")

####
# Interindividual variance is the variance of the random effects (random effects variance of 1|id)
# Residual variance of the random effects (within-individual variance)
### Example
# Groups   Name        Std.Dev.
 #id       (Intercept)  1.2 # interindividual variance
 #Residual              0.8 # within individual variance

# Extract random effects (BLUPs) for 'id1' from the model
random_effects <- ranef(out_rep$mod)$id1

# Calculate the variance of the random effects to estimate interindividual variance
interindividual_variance <- var(random_effects[, 1], na.rm = TRUE)
print(interindividual_variance)

# Extract residuals from the model
model_residuals <- residuals(out_rep$mod)

# Calculate the total residual variance (within-individual variance)
residual_variance <- var(model_residuals, na.rm = TRUE)
print(residual_variance)

# Calculate population-level repeatability (R)
R_population <- interindividual_variance / (interindividual_variance + residual_variance)
print(R_population)

library(dplyr)
#index of individual repeatabilitu
# Add residuals to the original data
# Extract residuals from the model
model_residuals <- residuals(out_rep$mod)
outbound$residuals <- model_residuals
mean_res <- mean(outbound$residuals, na.rm=TRUE)

# Calculate intraindividual variance for each individual (within-individual variance per individual)
intra_variance <- outbound %>%
  group_by(id1) %>%
  dplyr::summarise(S_ind =mean(residuals - mean_res,na.rm = TRUE)^2)

print(intra_variance)

# Calculate individual-level repeatability for each individual
intraindividual_repeatability <- intra_variance %>%
  mutate(R_ind = interindividual_variance / (interindividual_variance + S_ind))

# Print the individual-level repeatability values
print(intraindividual_repeatability)
mean(intraindividual_repeatability$R_ind, na.rm=TRUE)
sd(intraindividual_repeatability$R_ind, na.rm=TRUE)
```

Interindividual variance is estimated from the variance of the random effects for id1.
Within-individual variance is estimated from the variance of the residuals of the model.
Intraindividual variance is calculated as the variance of residuals for each individual.
Population-level repeatability and individual-level repeatability are calculated using these estimates


### Repeatability of the inbound commutes of February 2022 with Rpt

```{r}
inbound <- distsim_all %>%
  filter(direc=="inbound", sameid=="yes", year_cave=="dry 2022-1")%>%
  mutate(date=paste(date1, date2, sep="_"))

inbound$day_year1 <- yday(inbound$date1)
inbound$day_year2 <- yday(inbound$date2)
inbound$time_diff <- abs(inbound$day_year1-inbound$day_year2)


# Check the levels and counts
table(inbound$year_cave)

# Check for NAs in Distance, date, and id1
sum(is.na(inbound$Distance))
sum(is.na(inbound$date))
sum(is.na(inbound$id1))

# Convert id1 to factor if necessary
inbound$id1 <- as.factor(inbound$id1)

in_rep<- rpt(log(Distance)~ time_diff + (1|id1), grname = c("id1","Residual","Fixed"), 
    data = inbound, datatype = "Gaussian", nboot = 1000, npermut = 0, ratio = FALSE, adjusted = FALSE)

print(in_rep)

plot(in_rep, grname = "id1", scale = "probit")

#within-individual variance
0.094/(0.115+0.001)
0.039/(0.006+0.001)

#index of individual repeatabilitu
# Add residuals to the original data
# Extract residuals from the model
model_residuals <- residuals(in_rep$mod)
inbound$residuals <- model_residuals
mean_res <- mean(inbound$residuals, na.rm=TRUE)

# Calculate intraindividual variance for each individual (within-individual variance per individual)
intraindividual_variance <- inbound %>%
  group_by(id1) %>%
  dplyr::summarise(S_ind =mean(residuals - mean_res,na.rm = TRUE)^2)

print(intraindividual_variance)

# Calculate individual-level repeatability for each individual
intraindividual_repeatability <- intraindividual_variance %>%
  mutate(R_ind = interindividual_variance / (interindividual_variance + S_ind))

# Print the individual-level repeatability values
print(intraindividual_repeatability)
mean(intraindividual_repeatability$R_ind, na.rm=TRUE)
sd(intraindividual_repeatability$R_ind, na.rm=TRUE)
```

### Test how inbound and outbound commutes changes between tracking periods based on sameid

```{r}
# out_all <- distsim_all %>%
#   filter(direc=="outbound")

hist(log(distsim_all$Distance))
library(lmerTest)
m_path <- lmer(log(Distance) ~ year_cave*sameid*direc + (1|dyad), data = distsim_all)

summary(m_path)
plot(m_path)
car::Anova(m_path)

# Extract the model summary
model_summary <- summary(m_path)

# Extract t-values from the fixed effects
t_values <- model_summary$coefficients[, "t value"]

# Calculate p-values using the normal distribution
p_values <- 2 * (1 - pnorm(abs(t_values)))

# Create a table with the effect names, t-values, and p-values
effects_table <- data.frame(
  Effect = rownames(model_summary$coefficients),
  `t value` = t_values,
  `p value` = p_values
)

# Print the table
print(effects_table)

library(MuMIn)
r.squaredGLMM(m_path)

meanwithinbwteen <- distsim_all%>%
  group_by(sameid)%>%
  reframe(meandis=mean(Distance), sddist=sd(Distance))

meandirec <- distsim_all%>%
  group_by(direc)%>%
  reframe(meandis=mean(Distance), sddist=sd(Distance))


#random slope for id
# m_out_bp_rs <- lmer(log(Distance) ~ year_cave*sameid + (1|dyad), data = out_all)
# summary(m_out_bp_rs)
# 
# car::Anova(m_out_bp_rs)
# 
# simulationOutput <- simulateResiduals(fittedModel = m_out_bp_rs, plot = F)
# plot(simulationOutput)
# 
# AIC(m_out_bp, m_out_bp_rs)

#filter inbound commute
# in_all <- distsim_all %>%
#   filter(direc=="inbound")
# 
# m_in_bp <- lmer(log(Distance) ~ year_cave*sameid + (1|dyad), data = in_all)
# summary(m_in_bp)
# 
# car::Anova(m_in_bp)
# 
# simulationOutput <- simulateResiduals(fittedModel = m_in_bp, plot = F)
# plot(simulationOutput)
# 
# 
# m_in_bp_rs <- lmer(log(Distance) ~ year_cave+sameid + (1|dyad), data = in_all)
# summary(m_in_bp_rs)
# 
# car::Anova(m_in_bp_rs)
# 
# simulationOutput <- simulateResiduals(fittedModel = m_in_bp_rs, plot = F)
# plot(simulationOutput)
# 
# AIC(m_in_bp, m_in_bp_rs)

```

